### 2014-04-28 paper/paper.tex
initial

\documentclass[preprint]{sigplanconf}

\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{xypic}
\usepackage{multirow}

\usepackage{color}

\newcommand{\note}[1]{{\color{blue}{#1}}}

\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}

\newcommand{\play}{\mathsf{play}\;}
\newcommand{\sleep}{\mathsf{sleep}\;}
\newcommand{\sleepOp}{\textsf{sleep}}
\newcommand{\lang}{SonicPi}

\newcommand{\vtime}[1]{[#1]_{\mathsf{v}}}
\newcommand{\etime}[1]{[#1]_{\mathsf{t}}}

\newcommand{\interp}[1]{\llbracket{#1}\rrbracket}

\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}

\title{TIME MONAD}

\authorinfo{Sam Aaron}{}{}
\authorinfo{Dominic Orchard}{}{}
\title{A programming model for temporal coordination (in music)}

\begin{document}
\maketitle

\begin{figure}[t]
\[
\begin{array}{l}
\play C \\
\play E \\ 
\play G \\
\sleep 1 \\
\play F \\
\play A \\
\play C \\
\sleep 0.5 \\
\play G \\
\play B \\
\play D
\end{array}
\]
\caption{Playing three chords in \lang{} with the second two chords played
closer together by $0.5s$.}
\end{figure}

\paragraph{Terminology} 
We refer to closed (\ie{}, without free variables)
sequences of statements as \emph{programs}.

\begin{definition}
For a program $P$, then $\vtime{P}$ is the \emph{virtual time} that
elapses when running $P$ and $\etime{P}$ is the \emph{actual (kernel)
  time} that elapses when running $P$.
\end{definition}

\begin{definition}
\note{This is partial, there is no compositional definition for
$\etime{-}$ (proving this might be interesting, it can be
done by example).}
Since the \sleepOp{} operation increments virtual time, but also
causes kernel sleep then: $\vtime{\sleep t} = \etime{\sleep t} = t$,
and
%
\[
\vtime{P; Q} = \vtime{P} + \vtime{Q}
\]
\end{definition}

\begin{lemma}
For all programs $P$ then $\etime{P} \geq \vtime{P}$. 
\end{lemma}

\paragraph{Proof}
TODO

\begin{lemma}
For a program $P$ then:
%%
\begin{align*}
\etime{P; \sleep{} s} = 
 \begin{cases}
   s + \vtime{P} & \etime{P} \leq s \\
   \etime{P} & \etime{P} > s
 \end{cases}
\end{align*}
\end{lemma}

\begin{lemma}
For a program $P$ then
%%
\begin{align*}
\etime{\sleep{} s; P} = s + \etime{P}
\end{align*}
\end{lemma}

For example, consider subprograms $A$, $B$, $C$ interposed with two
sleep statements with lengths $s_1$ and $s_2$:
%
\begin{equation}
\begin{array}{l}
A \\
\sleep s_1 \\
B  \\
\sleep s_2 \\
C
\end{array}
\label{example:time1}
\end{equation}
%%
If $[A] = t_1$, $[B] = t_2$, $[C] = t_3$, then $[eq. \eqref{example:time1}] = 
s_1 + s_2 + t_3$, iff $t_1 \leq 1$ and $t_2 \leq 2$.

%\begin{equation*}
%\begin{array}{lllll}
%A & \multirow{2}{*}{\rule[1em]{0.6pt}{1.2em}} & \multirow{2}{*}{$t_1$} & 
%\multirow{4}{*}{\rule[1em]{0.6pt}{4em}} & \multirow{4}{*}{$t_1 + t_2 + 3$}
%\\
%\emph{sleep} \; 1 \qquad \\
%B &  \multirow{2}{*}{\rule[1em]{0.6pt}{1.2em}} & \multirow{2}{*}{$t_2$} \\
%\emph{sleep} \; 2
%\end{array}
%\end{equation*}

\subsection{Time monad}

\newcommand{\TM}{\mathsf{TM}}

\begin{equation*}
\TM a = (\mathit{Time}, \mathit{Time}) \rightarrow (\mathit{VTime} \rightarrow \mathit{IO} (a, \mathit{VTime}))
\end{equation*}

\paragraph{Quotienting by non-time dependent functions}

In L, there is no expression which returns the current time; 
 \emph{getTime} belongs only to the model, not to the language.
That is, for all expressions $e$, then the denotation 
$\interp{e}$ factors through

\subsection{Emitting overrun exceptions}

\paragraph{Overrun}

\paragraph{Overrun schedule}

\end{document}

### 2014-04-28 time-monad-alt.hs

@@ -1,0 +1,94 @@
+{-# LANGUAGE NoMonomorphismRestriction #-}
+
+import Data.Time
+import Control.Concurrent
+import System.IO.Unsafe
+
+import Debug.Trace
+
+data TimeMonad a = TM { unTM :: (UTCTime, UTCTime, NominalDiffTime) -> (a, NominalDiffTime) }
+
+instance Monad TimeMonad where
+    return a      = TM $ \(_, _, vt)  -> return (a, vt)
+    (TM c1) >>= f = TM $ \(startT, nowT, vT) -> let thenT = unsafePerformIO getCurrentTime
+                                                   (x, vT') = c1 (startT, thenT, vT)
+                                                in (unTM $ f x) (startT, nowT, vT')
+                                           
+
+runTime :: TimeMonad a -> IO a
+runTime (TM c) = do t <- getCurrentTime
+                    (y, _) <- c (t, t, 0)
+                    return y
+
+getTime :: TimeMonad UTCTime
+getTime = TM $ \(_, nowT, vT) -> return (nowT, vT)
+
+startTime :: TimeMonad UTCTime
+startTime = TM $ \(startT, _, vT) -> return (startT, vT)
+
+getVirtualTime :: TimeMonad NominalDiffTime
+getVirtualTime = TM $ \(_, _, vT) -> return (vT, vT)
+
+setVirtualTime :: NominalDiffTime -> TimeMonad ()
+setVirtualTime vT = TM $ \_ -> return ((), vT)
+
+sleep :: NominalDiffTime -> TimeMonad ()
+sleep delayT = do vT    <- getVirtualTime
+                  thenT <- startTime
+                  nowT  <- getTime
+                  let diffT = diffUTCTime nowT thenT
+                  let vT'   = vT + delayT
+                  if (vT' < diffT) then return () else kernelSleep (vT' - diffT)
+                  setVirtualTime vT'
+
+kernelSleep :: RealFrac a => a -> TimeMonad ()
+kernelSleep x = TM $ \(_, _, vT) -> do { threadDelay (round (x * 1000000)); return ((), vT) } 
+
+{- Testing -}
+
+foo = do a <- getTime
+         kernelSleep 2
+         sleep 2
+         b <- getTime
+         sleep 1
+         c <- getTime
+         return $ (diffUTCTime b a, -- 2
+                   diffUTCTime c b, -- 1
+                   diffUTCTime c a) -- 3
+
+foo2 = do a <- getTime
[…]

### 2014-04-28 time-monad.hs

@@ -1,0 +1,152 @@
+{-# LANGUAGE NoMonomorphismRestriction #-}
+
+import Data.Time
+import Control.Concurrent
+
+import Debug.Trace
+
+data TimeMonad a = TM { unTM :: (UTCTime, UTCTime, NominalDiffTime) -> IO (a, NominalDiffTime) }
+
+instance Monad TimeMonad where
+    return a      = TM $ \(_, _, vt)  -> return (a, vt)
+    (TM c1) >>= f = TM $ \(startT, nowT, vT) -> do thenT <- getCurrentTime
+                                                   (x, vT') <- c1 (startT, thenT, vT)
+                                                   (unTM $ f x) (startT, nowT, vT')
+                                           
+
+runTime :: TimeMonad a -> IO a
+runTime (TM c) = do t <- getCurrentTime
+                    (y, _) <- c (t, t, 0)
+                    return y
+
+time :: TimeMonad UTCTime
+time = TM $ \(_, nowT, vT) -> return (nowT, vT)
+
+startTime :: TimeMonad UTCTime
+startTime = TM $ \(startT, _, vT) -> return (startT, vT)
+
+virtualTime :: TimeMonad NominalDiffTime
+virtualTime = TM $ \(_, _, vT) -> return (vT, vT)
+
+setVirtualTime :: NominalDiffTime -> TimeMonad ()
+setVirtualTime vT = TM $ \_ -> return ((), vT)
+
+sleep :: NominalDiffTime -> TimeMonad ()
+sleep delayT = do vT    <- virtualTime
+                  thenT <- startTime
+                  nowT  <- time
+                  let diffT = diffUTCTime nowT thenT
+                  let vT'   = vT + delayT
+                  if (vT' < diffT) then return () else kernelSleep (vT' - diffT)
+                  setVirtualTime vT'
+
+kernelSleep :: RealFrac a => a -> TimeMonad ()
+kernelSleep x = TM $ \(_, _, vT) -> do { threadDelay (round (x * 1000000)); return ((), vT) } 
+
+{- Testing -}
+
+foo = do a <- time
+         kernelSleep 2
+         sleep 2
+         b <- time
+         sleep 1
+         c <- time
+         return $ (diffUTCTime b a, -- 2
+                   diffUTCTime c b, -- 1
+                   diffUTCTime c a) -- 3
+
+foo2 = do a <- time
+          sleep 1
+          b <- time
+          sleep 2
+          c <- time
+          return $ (diffUTCTime b a, 
[…]

### 2014-04-28 paper/paper.tex
some tweaks and paper textg

@@
 
+%% ODER: format ==         = "\mathrel{==}"
+%% ODER: format /=         = "\neq "
+%
+%
+\makeatletter
+\@ifundefined{lhs2tex.lhs2tex.sty.read}%
+  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
+   \newcommand\SkipToFmtEnd{}%
+   \newcommand\EndFmtInput{}%
+   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
+  }\SkipToFmtEnd
+
+\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
+\usepackage{amstext}
+\usepackage{amssymb}
+\usepackage{stmaryrd}
+\DeclareFontFamily{OT1}{cmtex}{}
+\DeclareFontShape{OT1}{cmtex}{m}{n}
+  {<5><6><7><8>cmtex8
+   <9>cmtex9
+   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
+\DeclareFontShape{OT1}{cmtex}{m}{it}
+  {<-> ssub * cmtt/m/it}{}
+\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
+\DeclareFontShape{OT1}{cmtt}{bx}{n}
+  {<5><6><7><8>cmtt8
+   <9>cmbtt9
+   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
+\DeclareFontShape{OT1}{cmtex}{bx}{n}
+  {<-> ssub * cmtt/bx/n}{}
+\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU
+
+\newcommand{\Sp}{\hskip.33334em\relax}
+
+\newcommand{\Conid}[1]{\mathit{#1}}
+\newcommand{\Varid}[1]{\mathit{#1}}
+\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
+\newcommand{\plus}{\mathbin{+\!\!\!+}}
+\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
+\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
+\newcommand{\sequ}{\mathbin{>\!\!\!>}}
+\renewcommand{\leq}{\leqslant}
+\renewcommand{\geq}{\geqslant}
+\usepackage{polytable}
+
+%mathindent has to be defined
+\@ifundefined{mathindent}%
+  {\newdimen\mathindent\mathindent\leftmargini}%
+  {}%
+
+\def\resethooks{%
+  \global\let\SaveRestoreHook\empty
+  \global\let\ColumnHook\empty}
+\newcommand*{\savecolumns}[1][default]%
+  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
+\newcommand*{\restorecolumns}[1][default]%
+  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
+\newcommand*{\aligncolumn}[2]%
+  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}
+
+\resethooks
+
+\newcommand{\onelinecommentchars}{\quad-{}- }
+\newcommand{\commentbeginchars}{\enskip\{-}
+\newcommand{\commentendchars}{-\}\enskip}
+
+\newcommand{\visiblecomments}{%
+  \let\onelinecomment=\onelinecommentchars
+  \let\commentbegin=\commentbeginchars
+  \let\commentend=\commentendchars}
+
+\newcommand{\invisiblecomments}{%
+  \let\onelinecomment=\empty
+  \let\commentbegin=\empty
+  \let\commentend=\empty}
+
+\visiblecomments
+
+\newlength{\blanklineskip}
+\setlength{\blanklineskip}{0.66084ex}
+
+\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
+\let\hspre\empty
+\let\hspost\empty
+\newcommand{\NB}{\textbf{NB}}
+\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}
+
+\EndFmtInput
+\makeatother
+%
+%
+%
+%
+%
+%
+% This package provides two environments suitable to take the place
+% of hscode, called "plainhscode" and "arrayhscode". 
+%
+% The plain environment surrounds each code block by vertical space,
+% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
+% similar to formulas. Note that if these dimensions are changed,
+% the spacing around displayed math formulas changes as well.
+% All code is indented using \leftskip.
+%
+% Changed 19.08.2004 to reflect changes in colorcode. Should work with
+% CodeGroup.sty.
+%
+\ReadOnlyOnce{polycode.fmt}%
+\makeatletter
+
+\newcommand{\hsnewpar}[1]%
+  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}
+
+% can be used, for instance, to redefine the code size, by setting the
+% command to \small or something alike
+\newcommand{\hscodestyle}{}
+
+% The command \sethscode can be used to switch the code formatting
+% behaviour by mapping the hscode environment in the subst directive
+% to a new LaTeX environment.
+
+\newcommand{\sethscode}[1]%
+  {\expandafter\let\expandafter\hscode\csname #1\endcsname
+   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}
+
+% "compatibility" mode restores the non-polycode.fmt layout.
+
+\newenvironment{compathscode}%
+  {\par\noindent
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \(\pboxed}%
+  {\endpboxed\)%
+   \par\noindent
+   \ignorespacesafterend}
+
+\newcommand{\compaths}{\sethscode{compathscode}}
+
+% "plain" mode is the proposed default.
+% It should now work with \centering.
+% This required some changes. The old version
+% is still available for reference as oldplainhscode.
+
+\newenvironment{plainhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\hspre\(\let\hspost\)%
+   \pboxed}%
+  {\endpboxed%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+\newenvironment{oldplainhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \(\pboxed}%
+  {\endpboxed\)%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+% Here, we make plainhscode the default environment.
+
+\newcommand{\plainhs}{\sethscode{plainhscode}}
+\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
+\plainhs
+
+% The arrayhscode is like plain, but makes use of polytable's
+% parray environment which disallows page breaks in code blocks.
+
+\newenvironment{arrayhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \(\parray}%
+  {\endparray\)%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+\newcommand{\arrayhs}{\sethscode{arrayhscode}}
+
+% The mathhscode environment also makes use of polytable's parray 
+% environment. It is supposed to be used only inside math mode 
+% (I used it to typeset the type rules in my thesis).
+
+\newenvironment{mathhscode}%
+  {\parray}{\endparray}
+
+\newcommand{\mathhs}{\sethscode{mathhscode}}
+
+% texths is similar to mathhs, but works in text mode.
+
+\newenvironment{texthscode}%
+  {\(\parray}{\endparray\)}
+
+\newcommand{\texths}{\sethscode{texthscode}}
+
+% The framed environment places code in a framed box.
+
+\def\codeframewidth{\arrayrulewidth}
+\RequirePackage{calc}
+
+\newenvironment{framedhscode}%
+  {\parskip=\abovedisplayskip\par\noindent
+   \hscodestyle
+   \arrayrulewidth=\codeframewidth
+   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
+   \hline\framedhslinecorrect\\{-1.5ex}%
+   \let\endoflinesave=\\
+   \let\\=\@normalcr
+   \(\pboxed}%
+  {\endpboxed\)%
+   \framedhslinecorrect\endoflinesave{.5ex}\hline
+   \endtabular
+   \parskip=\belowdisplayskip\par\noindent
+   \ignorespacesafterend}
+
+\newcommand{\framedhslinecorrect}[2]%
+  {#1[#2]}
+
+\newcommand{\framedhs}{\sethscode{framedhscode}}
+
+% The inlinehscode environment is an experimental environment
+% that can be used to typeset displayed code inline.
+
+\newenvironment{inlinehscode}%
+  {\(\def\column##1##2{}%
+   \let\>\undefined\let\<\undefined\let\\\undefined
+   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
+   \def\fromto##1##2##3{##3}%
+   \def\nextline{}}{\) }%
+
+\newcommand{\inlinehs}{\sethscode{inlinehscode}}
+
+% The joincode environment is a separate environment that
+% can be used to surround and thereby connect multiple code
+% blocks.
+
+\newenvironment{joincode}%
+  {\let\orighscode=\hscode
+   \let\origendhscode=\endhscode
+   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
+   %\let\SaveRestoreHook=\empty
+   %\let\ColumnHook=\empty
+   %\let\resethooks=\empty
+   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
+  {\origendhscode
+   \global\let\hscode=\orighscode
+   \global\let\endhscode=\origendhscode}%
+
+\makeatother
+\EndFmtInput
+%
+
+\usepackage{fancyvrb}
 \usepackage{stmaryrd}
@@
 \usepackage{multirow}
+\usepackage{hyperref}
+\usepackage{url}
+\usepackage{color}
 
-\usepackage{color}
+\ifdefined\nolhs
+\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small}
+\else
+\fi
 
@@
 \newcommand{\sleep}{\mathsf{sleep}\;}
+\newcommand{\playOp}{\textsf{play}}
 \newcommand{\sleepOp}{\textsf{sleep}}
@@
 \newcommand{\eg}{\emph{e.g.}}
-
-\title{TIME MONAD}
 
@@
 \end{figure}
+
+The \sleepOp{} statement essentially imposes a minimum \emph{soft deadline}
+(in the terminology of )
 
@@
 
+The semantics of time in the \lang{} language is monadic, that is,
+there is an underlying monad that describes the effectful component of
+the language, relating to time. In the following, we use Haskell as
+our meta language for the semantics (since it provides convenient
+syntax for working with monads)\footnote{The source code for the monadic
+model is avilable at \url{https://github.com/dorchard/time-monad}}.
+
 \newcommand{\TM}{\mathsf{TM}}
 
-\begin{equation*}
-\TM a = (\mathit{Time}, \mathit{Time}) \rightarrow (\mathit{VTime} \rightarrow \mathit{IO} (a, \mathit{VTime}))
-\end{equation*}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{data}\;\Conid{Temporal}\;\Varid{a}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{T}\;((\Conid{Time},\Conid{Time})\to (\Conid{VTime}\to \Conid{IO}\;(\Varid{a},\Conid{VTime}))){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%%
+Which maps a pair of two times, which will be the
+start time of the computation and current time, to
+a stateful computation over the \emph{IO} monad with
+a single location storing a virtual time.
+
+To model \lang{}, it is enough for \emph{Temporal} to have the
+structure of an \emph{applicative functor}.  Recall the usual
+applicative functor interface:
+%%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{class}\;\Conid{Functor}\;\Varid{f}\Rightarrow \Conid{Applicative}\;\Varid{f}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\Varid{pure}{}\<[10]%
+\>[10]{}\mathbin{::}\Varid{a}\to \Varid{f}\;\Varid{a}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}(<\!\!\!*\!\!\!>)\mathbin{::}\Varid{f}\;(\Varid{a}\to \Varid{b})\to \Varid{f}\;\Varid{a}\to \Varid{f}\;\Varid{b}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%%
+Thus, an instance for \emph{Temporal}
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{instance}\;\Conid{Applicative}\;\Conid{Temporal}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{pure}\;\Varid{a}{}\<[21]%
+\>[21]{}\mathrel{=}\Conid{T}\;(\lambda (\anonymous ,\anonymous )\to \lambda \Varid{vt}\to \Varid{return}\;(\Varid{a},\Varid{vt})){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}(\Conid{T}\;\Varid{f})<\!\!\!*\!\!\!>(\Conid{T}\;\Varid{x})\mathrel{=}{}\<[E]%
+\\
+\>[5]{}\hsindent{4}{}\<[9]%
+\>[9]{}\Conid{T}\;{}\<[12]%
+\>[12]{}(\lambda (\Varid{startT},\Varid{nowT})\to \lambda \Varid{vT}\to {}\<[E]%
+\\
+\>[12]{}\hsindent{3}{}\<[15]%
+\>[15]{}\mathbf{do}\;{}\<[19]%
+\>[19]{}\Varid{thenT}{}\<[30]%
+\>[30]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
+\\
+\>[19]{}(\Varid{x'},\Varid{vT'}){}\<[30]%
+\>[30]{}\leftarrow \Varid{x}\;(\Varid{startT},\Varid{thenT})\;\Varid{vT}{}\<[E]%
+\\
+\>[19]{}(\Varid{f'},\Varid{vT''})\leftarrow \Varid{f}\;(\Varid{startT},\Varid{nowT})\;\Varid{vT'}{}\<[E]%
+\\
+\>[19]{}\Varid{return}\;(\Varid{f'}\;\Varid{x'},\Varid{vT''})){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+\begin{align*}
+\interp{\emph{statement}} & : \emph{Temporal} \, () \\
+\interp{P; Q} & = (\lambda() \rightarrow \interp{P}) <\!\!\!*\!\!\!> \interp{Q} \\
+\interp{\sleep t} & = \emph{sleep} \, \interp{t}
+\end{align*}
+%%
+Note that $\interp{-}$ is overloaded in the rule for \sleepOp{} for (pure) expressions. 
+The concrete interpreation of other statements in the language, such as \playOp, is
+elided here since it does not relate directly to the temporal semantics.
+
+\begin{figure}[t]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{time},\Varid{start}\mathbin{::}\Conid{Temporal}\;\Conid{Time}{}\<[E]%
+\\
+\>[B]{}\Varid{time}{}\<[7]%
+\>[7]{}\mathrel{=}\Conid{T}\;(\lambda (\anonymous ,\Varid{nowT})\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{nowT},\Varid{vT})){}\<[E]%
+\\
+\>[B]{}\Varid{start}\mathrel{=}\Conid{T}\;(\lambda (\Varid{startT},\anonymous )\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{startT},\Varid{vT})){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{getVirtualTime}\mathbin{::}\Conid{Temporal}\;\Conid{VTime}{}\<[E]%
+\\
+\>[B]{}\Varid{getVirtualTime}\mathrel{=}\Conid{T}\;(\lambda (\anonymous ,\anonymous )\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{vT},\Varid{vT})){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{setVirtualTime}\mathbin{::}\Conid{VTime}\to \Conid{Temporal}\;(){}\<[E]%
+\\
+\>[B]{}\Varid{setVirtualTime}\;\Varid{vT}\mathrel{=}\Conid{T}\;(\lambda \anonymous \to \lambda \anonymous \to \Varid{return}\;((),\Varid{vT})){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\caption{Simple \emph{Temporal} computations, used  by the model}
+\end{figure}
+
+The key primitive \emph{sleep} provides the semantics for \sleepOp{} as:
+%%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{28}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{sleep}\mathbin{::}\Conid{VTime}\to \Conid{Temporal}\;(){}\<[E]%
+\\
+\>[B]{}\Varid{sleep}\;\Varid{delayT}\mathrel{=}\mathbf{do}\;{}\<[20]%
+\>[20]{}\Varid{vT}{}\<[28]%
+\>[28]{}\leftarrow \Varid{getVirtualTime}{}\<[E]%
+\\
+\>[20]{}\Varid{startT}{}\<[28]%
+\>[28]{}\leftarrow \Varid{start}{}\<[E]%
+\\
+\>[20]{}\Varid{nowT}{}\<[28]%
+\>[28]{}\leftarrow \Varid{time}{}\<[E]%
+\\
+\>[20]{}\mathbf{let}\;\Varid{diffT}\mathrel{=}\Varid{diffTime}\;\Varid{nowT}\;\Varid{startT}{}\<[E]%
+\\
+\>[20]{}\mathbf{let}\;\Varid{vT'}{}\<[30]%
+\>[30]{}\mathrel{=}\Varid{vT}\mathbin{+}\Varid{delayT}{}\<[E]%
+\\
+\>[20]{}\mathbf{if}\;(\Varid{vT'}\mathbin{<}\Varid{diffT}){}\<[E]%
+\\
+\>[20]{}\hsindent{2}{}\<[22]%
+\>[22]{}\mathbf{then}\;\Varid{return}\;(){}\<[E]%
+\\
+\>[20]{}\hsindent{2}{}\<[22]%
+\>[22]{}\mathbf{else}\;\Varid{kernelSleep}\;(\Varid{vT'}\mathbin{-}\Varid{diffT}){}\<[E]%
+\\
+\>[20]{}\Varid{setVirtualTime}\;\Varid{vT'}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 

### 2014-04-28 time-monad.hs

@@ -1,47 +1,63 @@
 {-# LANGUAGE NoMonomorphismRestriction #-}
 
 import Data.Time
+import Control.Applicative
 import Control.Concurrent
 
 import Debug.Trace
 
-data TimeMonad a = TM { unTM :: (UTCTime, UTCTime, NominalDiffTime) -> IO (a, NominalDiffTime) }
+type Time = UTCTime
+type VTime = NominalDiffTime
+diffTime = diffUTCTime
 
-instance Monad TimeMonad where
-    return a      = TM $ \(_, _, vt)  -> return (a, vt)
-    (TM c1) >>= f = TM $ \(startT, nowT, vT) -> do thenT <- getCurrentTime
-                                                   (x, vT') <- c1 (startT, thenT, vT)
-                                                   (unTM $ f x) (startT, nowT, vT')
+data Temporal a = T ((Time, Time) -> (VTime -> IO (a, VTime)))
+
+instance Functor Temporal where
+    fmap f (T x) = T (\(startT, nowT) -> \vT -> do (a, vT') <- x (startT, nowT) vT
+                                                   return (f a, vT'))
+
+instance Applicative Temporal where
+    pure a          = T (\(_, _) -> \vt -> return (a, vt))
+    (T f) <*> (T x) = T (\(startT, nowT) -> \vT -> 
+                             do thenT      <- getCurrentTime
+                                (x', vT')  <- x (startT, thenT) vT
+                                (f', vT'') <- f (startT, nowT) vT'
+                                return (f' x', vT''))
+
+instance Monad Temporal where
+    return a     = T $ \(_, _) -> \vt -> return (a, vt)
+    (T c1) >>= f = T $ \(startT, nowT) -> \vT -> do thenT <- getCurrentTime
+                                                    (x, vT') <- c1 (startT, thenT) vT
+                                                    let (T c2) = f x in c2 (startT, nowT) vT'
                                            
 
-runTime :: TimeMonad a -> IO a
-runTime (TM c) = do t <- getCurrentTime
-                    (y, _) <- c (t, t, 0)
-                    return y
+runTime :: Temporal a -> IO a
+runTime (T c) = do t <- getCurrentTime
+                   (y, _) <- c (t, t) 0
+                   return y
 
[…]

### 2014-04-28 paper/paper.tex
applicative versions and explanation

@@
 
+\bibliographystyle{amsalpha}
+
 \newcommand{\note}[1]{{\color{blue}{#1}}}
@@
 
+Thus ``$\sleep{} t$'' communicates that, after it has been evaluated, at least 
+$t$ seconds has elapsed since the last \sleepOp{}. This provides a kind of
+\emph{minimum (soft) deadline}.
+
+In \lang{}, it is possible that a computation proceeding a \sleepOp{}
+can overrun; that is, run longer than the sleep time.  Thus, the
+programming model is not suitable for realtime systems requiring hard
+deadlines but \sleepOp{} instead provides a \emph{soft deadline} (using
+the terminology of Hansson and Jonsson~\cite{hansson1994logic}).
+
 \paragraph{Terminology} 
@@
 For all programs $P$ then $\etime{P} \geq \vtime{P}$. 
+\label{lemma1}
+\end{lemma}
+
+\begin{lemma}
+For all programs $P$ and $Q$ then:
+%%
+\begin{equation}
+\vtime{P} + \vtime{Q} \leq \etime{P; Q} \leq \etime{P} + \etime{Q}
+\end{equation}
+\label{lemma2}
 \end{lemma}
@@
 \end{align*}
+\label{lemma3}
 \end{lemma}
@@
 \end{align*}
+\label{lemma4}
 \end{lemma}
@@
 
-\subsection{Time monad}
+\section{Model}
 
-The semantics of time in the \lang{} language is monadic, that is,
-there is an underlying monad that describes the effectful component of
-the language, relating to time. In the following, we use Haskell as
-our meta language for the semantics (since it provides convenient
-syntax for working with monads)\footnote{The source code for the monadic
-model is avilable at \url{https://github.com/dorchard/time-monad}}.
+In the following, we use Haskell as our meta language for the
+semantics (since it provides convenient syntax for working with
+monads)\footnote{The source code for the model is avilable at
+  \url{https://github.com/dorchard/time-monad}}.
 
@@
 a stateful computation over the \emph{IO} monad with
-a single location storing a virtual time.
+a single location storing a virtual time. The \emph{IO}
+computation provides underlying access to the actual
+time from kernel.
 
 To model \lang{}, it is enough for \emph{Temporal} to have the
-structure of an \emph{applicative functor}.  Recall the usual
-applicative functor interface:
+structure of an \emph{applicative functor}~\cite{mcbride2008functional} (also called
+\emph{idioms} or \emph{monoidal functors}).  Recall the usual
+applicative functor interface in Haskell:
 %%
@@
 %%
-Thus, an instance for \emph{Temporal}
+The \emph{Applicative} interface describes how to compose effectul
+computations encoded as values of type $f\ a$ (the effectful computation
+of a pure value of type $a$). Thus, \emph{pure} constructs a trivially
+effectful computation from a pure value. The \ensuremath{<\!\!\!*\!\!\!>} operation (which we'll
+call \emph{apply}) takes an effectful computation of a function and
+an effectful computation of an argument and evaluates the two effects
 
+The \emph{Applicative} instance for \emph{Temporal} is:
+%
 \begin{hscode}\SaveRestoreHook
@@
 \column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{21}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{30}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{44}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[21]{}\mathrel{=}\Conid{T}\;(\lambda (\anonymous ,\anonymous )\to \lambda \Varid{vt}\to \Varid{return}\;(\Varid{a},\Varid{vt})){}\<[E]%
-\\[\blanklineskip]%
+\\
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}(\Conid{T}\;\Varid{f})<\!\!\!*\!\!\!>(\Conid{T}\;\Varid{x})\mathrel{=}{}\<[E]%
+\>[5]{}(\Conid{T}\;\Varid{f})<\!\!\!*\!\!\!>(\Conid{T}\;\Varid{x})\mathrel{=}\Conid{T}\;(\lambda (\Varid{startT},\Varid{nowT})\to \lambda \Varid{vT}\to {}\<[E]%
 \\
-\>[5]{}\hsindent{4}{}\<[9]%
-\>[9]{}\Conid{T}\;{}\<[12]%
-\>[12]{}(\lambda (\Varid{startT},\Varid{nowT})\to \lambda \Varid{vT}\to {}\<[E]%
+\>[5]{}\hsindent{25}{}\<[30]%
+\>[30]{}\mathbf{do}\;\Varid{thenT}{}\<[44]%
+\>[44]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
 \\
-\>[12]{}\hsindent{3}{}\<[15]%
-\>[15]{}\mathbf{do}\;{}\<[19]%
-\>[19]{}\Varid{thenT}{}\<[30]%
-\>[30]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
+\>[30]{}\hsindent{3}{}\<[33]%
+\>[33]{}(\Varid{f'},\Varid{vT'}){}\<[44]%
+\>[44]{}\leftarrow \Varid{f}\;(\Varid{startT},\Varid{thenT})\;\Varid{vT}{}\<[E]%
 \\
-\>[19]{}(\Varid{x'},\Varid{vT'}){}\<[30]%
-\>[30]{}\leftarrow \Varid{x}\;(\Varid{startT},\Varid{thenT})\;\Varid{vT}{}\<[E]%
+\>[30]{}\hsindent{3}{}\<[33]%
+\>[33]{}(\Varid{x'},\Varid{vT''})\leftarrow \Varid{x}\;(\Varid{startT},\Varid{nowT})\;\Varid{vT'}{}\<[E]%
 \\
-\>[19]{}(\Varid{f'},\Varid{vT''})\leftarrow \Varid{f}\;(\Varid{startT},\Varid{nowT})\;\Varid{vT'}{}\<[E]%
-\\
-\>[19]{}\Varid{return}\;(\Varid{f'}\;\Varid{x'},\Varid{vT''})){}\<[E]%
+\>[30]{}\hsindent{3}{}\<[33]%
+\>[33]{}\Varid{return}\;(\Varid{f'}\;\Varid{x'},\Varid{vT''})){}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
+%
+The result of composing two temporal computations, with
+start time \emph{starT}, current time \emph{nowT}, and virtual
+time \emph{vT}, is the result of evaluating first the right 
+hand side, and then the left hand side
+
+Note this gives a left-to-right evaluation order on the appicative
+\emph{apply} operation.
+
+To understand this ordering, consider
 
@@
 \\
-\>[20]{}\mathbf{let}\;\Varid{diffT}\mathrel{=}\Varid{diffTime}\;\Varid{nowT}\;\Varid{startT}{}\<[E]%
-\\
 \>[20]{}\mathbf{let}\;\Varid{vT'}{}\<[30]%
 \>[30]{}\mathrel{=}\Varid{vT}\mathbin{+}\Varid{delayT}{}\<[E]%
+\\
+\>[20]{}\mathbf{let}\;\Varid{diffT}\mathrel{=}\Varid{diffTime}\;\Varid{nowT}\;\Varid{startT}{}\<[E]%
 \\
@@
 \end{hscode}\resethooks
+%%
+\emph{sleep} proceeds by calculating the elapsed time, \emph{diffT}, 
+and the new virtual time \emph{vT'}. If the new virtual time is less
+than the elapsed time then no actual (kernel) sleeping happens. However,
+if the new virtual time is ahead of the elapsed time, then the process
+sleeps for the difference. Finally, the virtual time state is updated
+with the new virtual time.
 
@@
 
+\section{Related work}
+
+There has been various work on reasoning about time in logic. For
+example, modal CTL (Computational Tree Logic) has been extended with
+time bounds for deadlines~\cite{emerson1991quantitative} (RCTL,
+Real-Time Computational Tree Logic) and for soft deadlines with
+probabilities on time bounds~\cite{hansson1994logic}. In these logics,
+temporal modalities are indexed with time bounds, \eg{}, $AF^{\leq 50}
+p$ means $p$ is true after at least $50$ ``time units'' (where $A$ is
+the CTL connective for \emph{along all paths} and $F$ for
+\emph{finally} (or \emph{eventually})). Our approach is less
+prescriptive and explicit, but has some resemblance in the use of
+\sleepOp{}. For example, the program $\sleep t ; P$ roughly
+corresponds to $AF^{\leq t} \interp{P}$, \ie{}, after at leat $t$ then
+whatever $P$ does will have happend. Our framework is not motivated by
+logic and we do not have a model checking process for answering
+questions such as, at time $t$ what formula hold (what statements have
+been evluated).  The properties of Lemmas 1 to 5 however provide some
+basis for programmers to reason about time in their programs. In
+practise, we find that such reasoning can be done by children in a
+completely informal but highly useful way.
+
+\section{Epilogue}
+
+\paragraph{Acknowledgements}
+
+\bibliography{references}
+
 \end{document}

### 2014-04-28 time-monad.hs

@@ -20,8 +20,8 @@ instance Applicative Temporal where
     pure a          = T (\(_, _) -> \vt -> return (a, vt))
     (T f) <*> (T x) = T (\(startT, nowT) -> \vT -> 
                              do thenT      <- getCurrentTime
-                                (x', vT')  <- x (startT, thenT) vT
-                                (f', vT'') <- f (startT, nowT) vT'
+                                (f', vT')  <- f (startT, thenT) vT
+                                (x', vT'') <- x (startT, nowT) vT'
                                 return (f' x', vT''))
 
 instance Monad Temporal where
@@ -61,6 +61,12 @@ kernelSleep x = T $ \(_, _) -> \vT -> do { threadDelay (round (x * 1000000)); re
 
 {- Testing -}
 
+timeTaken k = do k
+                 startT <- start
+                 endT <- time
+                 return $ (diffTime endT startT)
+
+
 foo = do a <- time
          kernelSleep 2
          sleep 2
@@ -86,11 +92,12 @@ foo3 = do sleep 1
           e <- time
           return $ (diffUTCTime e s)
 
+foo4app = (kernelSleep 2) *> (sleep 1) *> (sleep 2) *> (kernelSleep 3)
+
+
 foo4 = do kernelSleep 2
           sleep 1
           -- ;
-          --sleep 2
-          --kernelSleep 3
           foo4A
           s <- start
           e <- time

### 2014-04-28 paper/paper.tex
work on proofs and beginning examples

@@
 
+\usepackage{enumerate}
+\usepackage{subfigure}
 \usepackage{fancyvrb}
@@
 \newtheorem{lemma}{Lemma}
+\newtheorem{theorem}{Theorem}
 \theoremstyle{definition}
@@
 \newcommand{\play}{\mathsf{play}\;}
+\newcommand{\playOp}{\textsf{play}}
+
 \newcommand{\sleep}{\mathsf{sleep}\;}
-\newcommand{\playOp}{\textsf{play}}
 \newcommand{\sleepOp}{\textsf{sleep}}
+
+\newcommand{\ksleep}{\mathsf{kernelSleep}\;}
+\newcommand{\ksleepOp}{\textsf{kernelSleep}}
+
 \newcommand{\lang}{SonicPi}
@@
 
+\section{Introduction}
+
+\note{Introduction to SonicPi}
+
+The underlying programming model of SonicPi provides a way to separate
+the ordering of effects from the timing of
+effects. Figure~\ref{three-chord-example} shows an example program
+where three chords are played in sequence, combining simple notions of 
+parallel, timed, and ordered effects. The first three statements play
+the notes of a C major chord in parallel. A \sleepOp{} statement then 
+provides a \emph{barrier}
+
+Thus ``$\sleep{} t$'' communicates that, after it has been evaluated, at least 
+$t$ seconds has elapsed since the last \sleepOp{}. This provides a minimum
+time. In between calls to \sleepOp{}, any other statements can (with some limits)
+be considered task parallel.
+
+In \lang{}, it is possible that a computation proceeding a \sleepOp{}
+can overrun; that is, run longer than the sleep time.  Thus, the
+programming model is not suitable for realtime systems requiring hard
+deadlines but \sleepOp{} instead provides a \emph{soft deadline} (using
+the terminology of Hansson and Jonsson~\cite{hansson1994logic}).
+
+\begin{itemize}
+\item A computational model of virtual and real time 
+      structured by an \emph{applicative functor} abstraction.
+
+\item We show that the applicative approach extends
+      to a monadic approach, which can be composed with additional
+      monads to capture other useful notions of effect in \lang{} programs,
+      such as random numbers (Section~\ref{}).
+\end{itemize}
+
+Previously reported on the language~\cite{aaron2013sonic}
+
 \begin{figure}[t]
+\subfigure[Three chord program in \lang{}]{
+\begin{minipage}{0.46\linewidth}
 \[
+\hspace{-6em}
 \begin{array}{l}
@@
 \play B \\
-\play D
+\play D \\
 \end{array}
 \]
-\caption{Playing three chords in \lang{} with the second two chords played
+\end{minipage}
+\label{three-chord-example}
+}
+\subfigure[Timing of the three chord program]{
+\begin{minipage}{0.46\linewidth}
+\note{insert nice diagram that shows when the notes
+occur over the 1.5s duration} \\
+\end{minipage}
+\label{three-chord-timing}
+}
+\caption{Playing three chords (C major, F major, G major) 
+in \lang{} with the second two chords played
 closer together by $0.5s$.}
@@
 
-The \sleepOp{} statement essentially imposes a minimum \emph{soft deadline}
-(in the terminology of )
+\subsection{Examples}
+\label{sec:examples}
 
-Thus ``$\sleep{} t$'' communicates that, after it has been evaluated, at least 
-$t$ seconds has elapsed since the last \sleepOp{}. This provides a kind of
-\emph{minimum (soft) deadline}.
+\note{Show a few more example programs here that
+demonstrate the programming model.}
 
-In \lang{}, it is possible that a computation proceeding a \sleepOp{}
-can overrun; that is, run longer than the sleep time.  Thus, the
-programming model is not suitable for realtime systems requiring hard
-deadlines but \sleepOp{} instead provides a \emph{soft deadline} (using
-the terminology of Hansson and Jonsson~\cite{hansson1994logic}).
+Figure~\ref{sleep-examples} shows four similar programs which each
+have different internal behaviours for \sleepOp, illustrating the
+semantics of \sleepOp{}. The first three take 3s to execute and the
+last takes 4s to execute, with the behaviours:
+%
+\begin{enumerate}[(a)]
+\item{3s -- sleeps for 1s then sleeps for 2s (two sleeps performed);}
+\item{3s -- performs a computation lasting 1s, ignores
+the first \sleepOp{} since its minimum deadline has been reached, 
+and then sleeps for 2s (one sleep performed);}
+\item{3s -- performs a computation lasting 2s, which means that
+the first \sleepOp{} is ignored, and the second \sleepOp{} waits
+for only 1s to reach its minimum deadline (half a sleep performed);}
+\item{4s -- performs a computation lasting 2s, thus 
+the first \sleepOp{} is ignored, then performs a computation lasting
+2s, thus the second \sleepOp{} is ignored (no sleeps performed).}
+\end{enumerate}
 
-\paragraph{Terminology} 
-We refer to closed (\ie{}, without free variables)
-sequences of statements as \emph{programs}.
+\begin{figure}[t]
+\subfigure[Two sleeps]{
+\begin{minipage}{0.18\linewidth}
+\[
+\hspace{-1em}
+\begin{array}{l}
+\sleep 1 \\
+\sleep 2 \\ \\ \\ \\
+\end{array}
+\]
+\end{minipage}
+\label{sleep-examples:a}
+}
+\rule[-2em]{0.3pt}{5em}
+%\hspace{1em}
+% takes 3
+\subfigure[One sleep]{
+\begin{minipage}{0.23\linewidth}
+\begin{center}
+\[
+\hspace{-0.5em}
+\begin{array}{l}
+\ldots \; \textit{\# lasts 1s} \\
+\sleep 1 \\
+\sleep 2 \\ \\  \\
+\end{array}
+\]
+\end{center}
+\end{minipage}
+\label{sleep-examples:b}
+}
+\rule[-2em]{0.3pt}{5em}
+%\hspace{1em}
+% takes 3s
+\subfigure[Half a sleep]{
+\begin{minipage}{0.23\linewidth}
+\begin{center}
+\[
+\hspace{-0.5em}
+\begin{array}{l}
+\ldots \; \textit{\# lasts 2s} \\
+\sleep 1 \\
+\sleep 2 \\ \\ \\
+\end{array}
+\]
+\end{center}
+\end{minipage}
+\label{sleep-examples:c}
+% takes 6
+}
+\rule[-2em]{0.3pt}{5em}
+\subfigure[No sleeps]{
+\begin{minipage}{0.23\linewidth}
+\begin{center}
+\[
+\hspace{-0.5em}
+\begin{array}{l}
+\ldots \; \textit{\# lasts 2s} \\
+\sleep 1 \\
+\ldots \; \textit{\# lasts 2s} \\
+\sleep 2 \\  \\
+\end{array}
+\]
+\end{center}
+\end{minipage}
+}
 
+\caption{Example programs with different \sleepOp{} behaviours}
+\label{sleep-examples}
+\end{figure}
+
+\section{Model}
+
+\paragraph{Terminology and notation}
+We refer to closed sequences of statements (\ie{}, without free
+variables) as \emph{programs}. Throughout, $P$, $Q$ range over programs,
+and $s, t$ range over times (usually in seconds).
+
+\subsection{Virtual time and real time}
+
+The programming model of \lang{} distinguishes between the
+\emph{actual time} elapsed since the start of a program $P$, which we
+write as $\etime{P}$, and the \emph{virtual time} which is advanced by
+\sleepOp{} statements, written $\vtime{P}$.
+
+Since \sleepOp{} is the only operation that changes the virtual
+time, we can specification for $\vtime{-}$ over all programs:
+%
 \begin{definition}
-For a program $P$, then $\vtime{P}$ is the \emph{virtual time} that
-elapses when running $P$ and $\etime{P}$ is the \emph{actual (kernel)
-  time} that elapses when running $P$.
+Virtual time is specified for statements of \lang{} programs 
+by the following (ordered) cases:
+%
+\begin{align*}
+\vtime{\sleep t} & = t \\ 
+\vtime{P; Q} & = \vtime{P} + \vtime{Q} \\
+\vtime{-} & = 0
+\end{align*}
+%
+\ie{}, the virtual time is $0$ 
+for any statment other than \sleepOp{} or sequential composition.
+\label{sleep-spec}
 \end{definition}
 
-\begin{definition}
-\note{This is partial, there is no compositional definition for
-$\etime{-}$ (proving this might be interesting, it can be
-done by example).}
-Since the \sleepOp{} operation increments virtual time, but also
-causes kernel sleep then: $\vtime{\sleep t} = \etime{\sleep t} = t$,
-and
-%
-\[
-\vtime{P; Q} = \vtime{P} + \vtime{Q}
-\]
-\end{definition}
+Equality of computation times is difficult. In order to not sweep this
+issue entirely under the carpet, we will use the relation $\approx$,
+where $s \approx t$ means that $s$ and $t$ are either exactly equal or
+differ by some amount $\epsilon$.\note{Discuss this further, may be
+  able to say later that in some cases $\epsilon$ is the scheduling
+  time for play statments?}  The virtual time and actual time of a
+single sleep statement (in isolation) are roughly the same, \ie{},
+$\vtime{\sleep t} \approx \etime{\sleep t}$ and thus $\vtime{\sleep t}
+\approx t$ by the specification in Definition~\ref{sleep-spec}.
+
+We should be clear however, that this approximate equality does not always
+hold. We can only say that it holds when \sleepOp{} is used in isolation, that is, 
+when it is the only statement in a program. As shown by the examples
+of Section~\ref{sec:examples}, the use of $\sleep t$ in a program 
+does not mean that a program necessarily waits for $t$ seconds-- 
+depending on the context, it may wait for anywhere between $0$ and $t$ seconds.
+
+For convenience, and to contrast with \sleepOp{}, we'll use an additional
+statement \ksleepOp{} here (which is not available in the actual language)
+ which always sleeps for the number of seconds specified by its parameter.
 
 \begin{lemma}
-For all programs $P$ then $\etime{P} \geq \vtime{P}$. 
-\label{lemma1}
-\end{lemma}
-
-\begin{lemma}
-For all programs $P$ and $Q$ then:
-%%
-\begin{equation}
-\vtime{P} + \vtime{Q} \leq \etime{P; Q} \leq \etime{P} + \etime{Q}
-\end{equation}
-\label{lemma2}
-\end{lemma}
-
-\paragraph{Proof}
-TODO
-
-\begin{lemma}
-For a program $P$ then:
+For some program $P$ and time $s$: 
 %%
@@
 \end{align*}
-\label{lemma3}
+\label{lem:sleep-R}
 \end{lemma}
@@
 \begin{lemma}
-For a program $P$ then
+For some program $P$ and time $s$:
 %%
@@
 \end{align*}
-\label{lemma4}
+\label{lem:sleep-L}
 \end{lemma}
+
+\begin{lemma}
+For all programs $P$ then $\etime{P} \geq \vtime{P}$. 
+\label{lemma1}
+\end{lemma}
+
+\begin{theorem}
+For all programs $P$ and $Q$ then:
+%%
+\begin{equation}
+\vtime{P} + \vtime{Q} \leq \etime{P; Q} \leq \etime{P} + \etime{Q}
+\end{equation}
+\label{len:sleep}
+\end{theorem}
+
+\paragraph{Proof}
+Since sequential composition is associative, we can reassociate
+$P; Q$ such that $P = P_1; P'$ where $P_1$ is a single statement
+
+\begin{itemize}
+\item $P = \sleep t$
+
+\begin{align*}
+\begin{array}{ll}
+       & \etime{\sleep t; Q} \\[0.5em]
+\equiv & \; \{\textit{Lemma}~\ref{lem:sleep-L}\} \\[0.1em]
+       & s + \etime{Q} \\[0.5em]
+\equiv & \; \{\textit{Definition}~\ref{}\} \\[0.1em]
+       & \etime{\sleep t} + \etime{Q}
+\end{array}
+\end{align*}
+
+\item $Q = \sleep t$
+
+\begin{itemize}
+\item $\etime{P} \leq t$ then:
+
+\begin{align*}
+\begin{array}{ll}
+       & \etime{P; \sleep t} \\[0.5em]
+\equiv & \; \{\textit{Lemma}~\ref{lem:sleep-R}\} \\[0.1em]
+       & \vtime{P} + t \\[0.5em]
+\equiv & \; \{\textit{Definition}~\ref{sleep-spec}\} \\[0.1em]
+       & \vtime{P} + \vtime{\sleep t}
+\end{array}
+\end{align*}
+
+ $\etime{P; \sleep t} = \etime{P} + $
+
+\end{itemize}
+
+\item $P = P';P'', Q = Q';Q''$
+
+By induction:
+%%
+\begin{align}
+\vtime{P'} + \vtime{P''} \leq \etime{P'; P''} \leq \vtime{P'} + \vtime{P''} \\
+\vtime{Q'} + \vtime{Q''} \leq \etime{Q'; Q''} \leq \vtime{Q'} + \vtime{Q''} \\
+\end{align}
+
+and
+
+\end{itemize}
 
@@
 
-\section{Model}
+\subsection{Applicative structure on computation}
 
@@
 \column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{28}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{time},\Varid{start}\mathbin{::}\Conid{Temporal}\;\Conid{Time}{}\<[E]%
+\>[B]{}\Varid{time}\mathbin{::}\Conid{Temporal}\;\Conid{Time}{}\<[E]%
 \\
@@
 \>[7]{}\mathrel{=}\Conid{T}\;(\lambda (\anonymous ,\Varid{nowT})\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{nowT},\Varid{vT})){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{start}\mathbin{::}\Conid{Temporal}\;\Conid{Time}{}\<[E]%
 \\
@@
 \>[B]{}\Varid{setVirtualTime}\;\Varid{vT}\mathrel{=}\Conid{T}\;(\lambda \anonymous \to \lambda \anonymous \to \Varid{return}\;((),\Varid{vT})){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{kernelSleep}\mathbin{::}\Conid{RealFrac}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{Temporal}\;(){}\<[E]%
+\\
+\>[B]{}\Varid{kernelSleep}\;\Varid{t}\mathrel{=}{}\<[18]%
+\>[18]{}\Conid{T}\;(\lambda (\anonymous ,\anonymous )\to \lambda \Varid{vT}\to {}\<[E]%
+\\
+\>[18]{}\hsindent{6}{}\<[24]%
+\>[24]{}\mathbf{do}\;{}\<[28]%
+\>[28]{}\Varid{threadDelay}\;(\Varid{round}\;(\Varid{t}\mathbin{*}\mathrm{1000000})){}\<[E]%
+\\
+\>[28]{}\Varid{return}\;((),\Varid{vT})){}\<[E]%
 \ColumnHook
@@
 basis for programmers to reason about time in their programs. In
-practise, we find that such reasoning can be done by children in a
+practice, we find that such reasoning can be done by children in a
 completely informal but highly useful way.

### 2014-04-28 time-monad.hs

@@ -66,6 +66,11 @@ timeTaken k = do k
                  endT <- time
                  return $ (diffTime endT startT)
 
+fig2a = (sleep 1) *> (sleep 2)
+fig2b = (kernelSleep 1) *> (sleep 1) *> (sleep 2)
+fig2c = (kernelSleep 2) *> (sleep 1) *> (sleep 2)
+fig2d = (kernelSleep 2) *> (sleep 1) *> (kernelSleep 2) *> (sleep 2) 
+
 
 foo = do a <- time
          kernelSleep 2
@@ -91,6 +96,9 @@ foo3 = do sleep 1
           s <- start
           e <- time
           return $ (diffUTCTime e s)
+
+
+
 
 foo4app = (kernelSleep 2) *> (sleep 1) *> (sleep 2) *> (kernelSleep 3)
 

### 2014-04-29 paper/paper.tex
latest stuff on proofs and model

@@
 \section{Introduction}
+\label{sec:introduction}
 
@@
 \begin{lemma}
-For some program $P$ and time $s$: 
+For some program $P$ and time $t$: 
 %%
 \begin{align*}
-\etime{P; \sleep{} s} = 
+\etime{P; \sleep{} t} = 
  \begin{cases}
-   s + \vtime{P} & \etime{P} \leq s \\
-   \etime{P} & \etime{P} > s
+   \etime{P} & (\vtime{P} + t) < \etime{P} \\
+   \vtime{P} + t  & \textit{otherwise}
  \end{cases}
 \end{align*}
+%\begin{align*}
+%\etime{P; \sleep{} t} = 
+% \begin{cases}
+%   \etime{P} & t < \etime{P} \\
+%   \vtime{P} + t & t \geq \etime{P}
+% \end{cases}
+%\end{align*}
 \label{lem:sleep-R}
@@
 \begin{lemma}
-For some program $P$ and time $s$:
+For some program $P$ and time $t$:
 %%
 \begin{align*}
-\etime{\sleep{} s; P} = s + \etime{P}
+\etime{\sleep{} t; P} = t + \etime{P}
 \end{align*}
@@
 \end{equation}
-\label{len:sleep}
+\label{theorem:main}
 \end{theorem}
 
-\paragraph{Proof}
-Since sequential composition is associative, we can reassociate
-$P; Q$ such that $P = P_1; P'$ where $P_1$ is a single statement
-
-\begin{itemize}
-\item $P = \sleep t$
-
-\begin{align*}
-\begin{array}{ll}
-       & \etime{\sleep t; Q} \\[0.5em]
-\equiv & \; \{\textit{Lemma}~\ref{lem:sleep-L}\} \\[0.1em]
-       & s + \etime{Q} \\[0.5em]
-\equiv & \; \{\textit{Definition}~\ref{}\} \\[0.1em]
-       & \etime{\sleep t} + \etime{Q}
-\end{array}
-\end{align*}
-
-\item $Q = \sleep t$
-
-\begin{itemize}
-\item $\etime{P} \leq t$ then:
-
-\begin{align*}
-\begin{array}{ll}
-       & \etime{P; \sleep t} \\[0.5em]
-\equiv & \; \{\textit{Lemma}~\ref{lem:sleep-R}\} \\[0.1em]
-       & \vtime{P} + t \\[0.5em]
-\equiv & \; \{\textit{Definition}~\ref{sleep-spec}\} \\[0.1em]
-       & \vtime{P} + \vtime{\sleep t}
-\end{array}
-\end{align*}
-
- $\etime{P; \sleep t} = \etime{P} + $
-
-\end{itemize}
-
-\item $P = P';P'', Q = Q';Q''$
-
-By induction:
-%%
-\begin{align}
-\vtime{P'} + \vtime{P''} \leq \etime{P'; P''} \leq \vtime{P'} + \vtime{P''} \\
-\vtime{Q'} + \vtime{Q''} \leq \etime{Q'; Q''} \leq \vtime{Q'} + \vtime{Q''} \\
-\end{align}
-
-and
-
-\end{itemize}
-
-For example, consider subprograms $A$, $B$, $C$ interposed with two
-sleep statements with lengths $s_1$ and $s_2$:
+From these lemmas we can reason about the evaluation time of programs.
+For example, consider subprograms $A$, $B$, $C$ where 
+$\etime{A} = t_1$, $\etime{B} = t_2$, $\etime{C} = t_3$ and
+$\vtime{A} = \vtime{B} \vtime{C} = 0$, interposed with two
+sleep statements of duration $s_1$ and $s_2$:
 %
@@
 %%
-If $[A] = t_1$, $[B] = t_2$, $[C] = t_3$, then $[eq. \eqref{example:time1}] = 
+Then by the above lemmas, we see that $\etime{eq. \eqref{example:time1}} = 
 s_1 + s_2 + t_3$, iff $t_1 \leq 1$ and $t_2 \leq 2$.
@@
 
-\subsection{Applicative structure on computation}
+\newcommand{\TM}{\mathsf{TM}}
+
+\subsection{Monadic structure on computation}
 
@@
   \url{https://github.com/dorchard/time-monad}}.
-
-\newcommand{\TM}{\mathsf{TM}}
-
+\lang{} computations are modelled by the \emph{Temporal} data type, defined:
+%%
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-%%
-Which maps a pair of two times, which will be the
-start time of the computation and current time, to
-a stateful computation over the \emph{IO} monad with
-a single location storing a virtual time. The \emph{IO}
-computation provides underlying access to the actual
-time from kernel.
-
-To model \lang{}, it is enough for \emph{Temporal} to have the
-structure of an \emph{applicative functor}~\cite{mcbride2008functional} (also called
-\emph{idioms} or \emph{monoidal functors}).  Recall the usual
-applicative functor interface in Haskell:
-%%
+%
+Thus, temporal computations map a pair of two times, which will be
+the start time of the computation and current time, to a stateful
+computation with a single location storing a virtual time, over the
+\emph{IO} type.  The \emph{IO} computation provides underlying access
+to the actual time from kernel.
+
+The \emph{Monad} instance for \emph{Temporal} is then as follows:
+%
 \begin{hscode}\SaveRestoreHook
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{37}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{class}\;\Conid{Functor}\;\Varid{f}\Rightarrow \Conid{Applicative}\;\Varid{f}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}\Varid{pure}{}\<[10]%
-\>[10]{}\mathbin{::}\Varid{a}\to \Varid{f}\;\Varid{a}{}\<[E]%
-\\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}(<\!\!\!*\!\!\!>)\mathbin{::}\Varid{f}\;(\Varid{a}\to \Varid{b})\to \Varid{f}\;\Varid{a}\to \Varid{f}\;\Varid{b}{}\<[E]%
+\>[B]{}\mathbf{instance}\;\Conid{Monad}\;\Conid{Temporal}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{return}\;\Varid{a}{}\<[16]%
+\>[16]{}\mathrel{=}\Conid{T}\;(\lambda \anonymous \to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{a},\Varid{vT})){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\\[-1.5em]{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}(\Conid{T}\;\Varid{p})\bind \Varid{q}{}\<[16]%
+\>[16]{}\mathrel{=}\Conid{T}\;(\lambda (\Varid{startT},\Varid{nowT})\to \lambda \Varid{vT}\to {}\<[E]%
+\\
+\>[16]{}\hsindent{5}{}\<[21]%
+\>[21]{}\mathbf{do}\;{}\<[25]%
+\>[25]{}(\Varid{x},\Varid{vT'}){}\<[37]%
+\>[37]{}\leftarrow \Varid{p}\;(\Varid{startT},\Varid{nowT})\;\Varid{vT}{}\<[E]%
+\\
+\>[25]{}\mathbf{let}\;(\Conid{T}\;\Varid{q'}){}\<[37]%
+\>[37]{}\mathrel{=}\Varid{q}\;\Varid{x}{}\<[E]%
+\\
+\>[25]{}\Varid{thenT}{}\<[37]%
+\>[37]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
+\\
+\>[25]{}\Varid{q'}\;(\Varid{startT},\Varid{thenT})\;\Varid{vT'}){}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
-%%
-The \emph{Applicative} interface describes how to compose effectul
-computations encoded as values of type $f\ a$ (the effectful computation
-of a pure value of type $a$). Thus, \emph{pure} constructs a trivially
-effectful computation from a pure value. The \ensuremath{<\!\!\!*\!\!\!>} operation (which we'll
-call \emph{apply}) takes an effectful computation of a function and
-an effectful computation of an argument and evaluates the two effects
-
-The \emph{Applicative} instance for \emph{Temporal} is:
-%
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{44}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{instance}\;\Conid{Applicative}\;\Conid{Temporal}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{pure}\;\Varid{a}{}\<[21]%
-\>[21]{}\mathrel{=}\Conid{T}\;(\lambda (\anonymous ,\anonymous )\to \lambda \Varid{vt}\to \Varid{return}\;(\Varid{a},\Varid{vt})){}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}(\Conid{T}\;\Varid{f})<\!\!\!*\!\!\!>(\Conid{T}\;\Varid{x})\mathrel{=}\Conid{T}\;(\lambda (\Varid{startT},\Varid{nowT})\to \lambda \Varid{vT}\to {}\<[E]%
-\\
-\>[5]{}\hsindent{25}{}\<[30]%
-\>[30]{}\mathbf{do}\;\Varid{thenT}{}\<[44]%
-\>[44]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
-\\
-\>[30]{}\hsindent{3}{}\<[33]%
-\>[33]{}(\Varid{f'},\Varid{vT'}){}\<[44]%
-\>[44]{}\leftarrow \Varid{f}\;(\Varid{startT},\Varid{thenT})\;\Varid{vT}{}\<[E]%
-\\
-\>[30]{}\hsindent{3}{}\<[33]%
-\>[33]{}(\Varid{x'},\Varid{vT''})\leftarrow \Varid{x}\;(\Varid{startT},\Varid{nowT})\;\Varid{vT'}{}\<[E]%
-\\
-\>[30]{}\hsindent{3}{}\<[33]%
-\>[33]{}\Varid{return}\;(\Varid{f'}\;\Varid{x'},\Varid{vT''})){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-%
+%
+\begin{itemize}
+
+\item \ensuremath{\Varid{return}\mathbin{::}\Varid{a}\to \Conid{Temporal}\;\Varid{a}} thus ignores the time parameters and
+has the usual pure state behaviour: passing the state unchanged.
+For the bind
+
+\item \ensuremath{(\bind )\mathbin{::}\Conid{Temporal}\;\Varid{a}\to (\Varid{a}\to \Conid{Temporal}\;\Varid{b})\to \Conid{Temporal}\;\Varid{b}} thus
+composes two computations together. 
 The result of composing two temporal computations, with
@@
 
-Note this gives a left-to-right evaluation order on the appicative
-\emph{apply} operation.
-
-To understand this ordering, consider
-
-\begin{align*}
-\interp{\emph{statement}} & : \emph{Temporal} \, () \\
-\interp{P; Q} & = (\lambda() \rightarrow \interp{P}) <\!\!\!*\!\!\!> \interp{Q} \\
-\interp{\sleep t} & = \emph{sleep} \, \interp{t}
-\end{align*}
-%%
-Note that $\interp{-}$ is overloaded in the rule for \sleepOp{} for (pure) expressions. 
-The concrete interpreation of other statements in the language, such as \playOp, is
-elided here since it does not relate directly to the temporal semantics.
+\end{itemize}
+
+The following \ensuremath{\Varid{runTime}} operation executes a temporal computation
+inside of the \emph{IO} monad by providing the start time of the
+computation, and with virtual time 0:
+%%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{runTime}\mathbin{::}\Conid{Temporal}\;\Varid{a}\to \Conid{IO}\;\Varid{a}{}\<[E]%
+\\
+\>[B]{}\Varid{runTime}\;(\Conid{T}\;\Varid{c})\mathrel{=}\mathbf{do}\;{}\<[21]%
+\>[21]{}\Varid{startT}\leftarrow \Varid{getCurrentTime}{}\<[E]%
+\\
+\>[21]{}(\Varid{x},\anonymous )\leftarrow \Varid{c}\;(\Varid{startT},\Varid{startT})\;\mathrm{0}{}\<[E]%
+\\
+\>[21]{}\Varid{return}\;\Varid{x}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%%
+To illustrate the evalution of temporal computation and the
+ordering and interleaving of calls to the operation system for the
+current time, we consider a program:
+%%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{runTime}\;(\mathbf{do}\;\{\mskip1.5mu \Varid{f};\Varid{g};\Varid{h};\mskip1.5mu\}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+(where \ensuremath{\Varid{f}\mathrel{=}\Conid{T}\;\Varid{f'},\Varid{g}\mathrel{=}\Conid{T}\;\Varid{g'},\Varid{h}\mathrel{=}\Conid{T}\;\Varid{h'}}) which 
+desugars to the following, after some simplification: 
+%%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{do}\;{}\<[5]%
+\>[5]{}\Varid{startT}{}\<[15]%
+\>[15]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
+\\
+\>[5]{}(\anonymous ,\Varid{vT'}){}\<[15]%
+\>[15]{}\leftarrow \Varid{f'}\;(\Varid{startT},\Varid{startT})\;\mathrm{0}{}\<[E]%
+\\
+\>[5]{}\Varid{thenT}{}\<[15]%
+\>[15]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
+\\
+\>[5]{}(\anonymous ,\Varid{vT''})\leftarrow \Varid{g'}\;(\Varid{startT},\Varid{thenT})\;\Varid{vT'}{}\<[E]%
+\\
+\>[5]{}\Varid{thenT'}{}\<[15]%
+\>[15]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
+\\
+\>[5]{}(\Varid{y},\anonymous ){}\<[15]%
+\>[15]{}\leftarrow \Varid{h'}\;(\Varid{startT},\Varid{thenT'})\;\Varid{vT''})\;{}\<[E]%
+\\
+\>[5]{}\Varid{return}\;\Varid{y}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%
+The above illustrates the repeated calls to \ensuremath{\Varid{getCurrentTime}}, the
+constant start time parameter, and the threading of virtual time state
+throughout the computation.
 
@@
 
+\paragraph{Interpreting \lang{} statements}
+
+The following interpretation function $\interp{-}$ on \lang{} 
+programs shows the mapping to the operations of the \emph{Temporal}
+monad:
+%%
+\begin{align*}
+\interp{\emph{statement}} & : \emph{Temporal} \, () \\
+\interp{x = P; Q} & = \interp{P} \ensuremath{\bind (\lambda \Varid{x}\to } \interp{Q}) \\
+\interp{P; Q} & = \interp{P} \ensuremath{\bind (\lambda \anonymous \to } \interp{Q}) \\
+\interp{\sleep e} & = \emph{sleep} \, \interp{e}
+\end{align*}
+%%
+Note that $\interp{-}$ is overloaded in the rule for \sleepOp{} for (pure) expressions. 
+The concrete interpreation of other statements in the language, such as \playOp, is
+elided here since it does not relate directly to the temporal semantics.
+
 The key primitive \emph{sleep} provides the semantics for \sleepOp{} as:
@@
 %%
-\emph{sleep} proceeds by calculating the elapsed time, \emph{diffT}, 
-and the new virtual time \emph{vT'}. If the new virtual time is less
-than the elapsed time then no actual (kernel) sleeping happens. However,
-if the new virtual time is ahead of the elapsed time, then the process
-sleeps for the difference. Finally, the virtual time state is updated
+Thus, \emph{sleep} proceeds by calculating the elapsed time, \emph{diffT}, 
+and the new virtual time \emph{vT'} (after the sleep). 
+If the new virtual time is less than the elapsed time then no actual (kernel) 
+sleeping happens. However, if the new virtual time is ahead of the elapsed time, 
+then the process sleeps for the difference. Finally, the virtual time state is updated
 with the new virtual time.
 
+\subsection{Soundness}
+
+\paragraph{Lemma~\ref{lem:sleep-R}}
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{do}\;\Varid{startT}\leftarrow \Varid{getCurrentTime}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}(\Varid{x},\Varid{vT'}){}\<[16]%
+\>[16]{}\leftarrow \Varid{p}\;(\Varid{startT},\Varid{startT})\;\mathrm{0}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\Varid{thenT}{}\<[16]%
+\>[16]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\mathbf{let}\;\Varid{vT''}{}\<[15]%
+\>[15]{}\mathrel{=}\Varid{vT'}\mathbin{+}\Varid{t}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\mathbf{let}\;\Varid{diffT}\mathrel{=}\Varid{diffTime}\;\Varid{thenT}\;\Varid{startT}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\mathbf{if}\;(\Varid{vT''}\mathbin{<}\Varid{diffT}){}\<[E]%
+\\
+\>[4]{}\hsindent{3}{}\<[7]%
+\>[7]{}\mathbf{then}\;\Varid{return}\;(){}\<[E]%
+\\
+\>[4]{}\hsindent{3}{}\<[7]%
+\>[7]{}\mathbf{else}\;\Varid{kernelSleep}\;(\Varid{vT''}\mathbin{-}\Varid{diffT}){}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\Varid{setVirtualTime}\;\Varid{vT''}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+\begin{align*}
+\etime{P; \sleep{} t} = 
+ \begin{cases}
+   \etime{P} & (\vtime{P} + t) < \etime{P}  \\
+   \etime{P} + (\vtime{P} + t) - \etime{P}  & \textit{otherwise}
+ \end{cases}
+\end{align*}
+
+\begin{align*}
+\etime{P; \sleep{} t} = 
+ \begin{cases}
+   \etime{P} & (\vtime{P} + t) < \etime{P} \\
+   \vtime{P} + t  &  \textit{otherwise}
+ \end{cases}
+\end{align*}
+
+\paragraph{Lemma~\ref{lem:sleep-L}}
+ $\etime{\sleep{} t; P} = t + \etime{P}$.
+This lemma holds in the model, where \ensuremath{\Varid{runTime}\;(\mathbf{do}\;\{\mskip1.5mu \Varid{sleep}\;\Varid{t};\Conid{P}\mskip1.5mu\})} desugars and
+simplifies to:
+%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{do}\;{}\<[5]%
+\>[5]{}\Varid{startT}\leftarrow \Varid{getCurrentTime}{}\<[E]%
+\\
+\>[5]{}\Varid{kernelSleep}\;\Varid{t}{}\<[E]%
+\\
+\>[5]{}\mathbf{let}\;(\Conid{T}\;\Varid{p'}){}\<[17]%
+\>[17]{}\mathrel{=}\Varid{p}\;(){}\<[E]%
+\\
+\>[5]{}\Varid{thenT}{}\<[12]%
+\>[12]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
+\\
+\>[5]{}\Varid{p'}\;(\Varid{startT},\Varid{thenT})\;\Varid{t}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%
+Thus, we see that we \ensuremath{\Varid{kernelSleep}} for $t$, and then continue with $P$ at virtual
+time $t$.
+
 \paragraph{Quotienting by non-time dependent functions}
+
+\note{This section is more of a marker for myself (Dom), I need
+to think about this later, but basically it is about showing
+that the monad laws hold for our semantics (but not in general)}
 
@@
 $\interp{e}$ factors through
+
+\subsection{Subsets of the semantics}
+
+For the examples of Section~\ref{sec:introduction}, the full structure
+of monad is not needed to give their semantics as there is no using of
+binding between statements (and thus no dataflow). In these case just
+an \emph{applicative functor}~\cite{mcbride2008functional} or even a
+monoid would suffice. These can be derived from the monad structure
+on \emph{Temporal} since all monads are applicative functors and all 
+monads define monoid over $m ()$.
+
+\paragraph{Applicative subset}
+
+Applicative functors are described by the following interface in
+Haskell:
+%%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{class}\;\Conid{Functor}\;\Varid{f}\Rightarrow \Conid{Applicative}\;\Varid{f}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\Varid{pure}{}\<[10]%
+\>[10]{}\mathbin{::}\Varid{a}\to \Varid{f}\;\Varid{a}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}(<\!\!\!*\!\!\!>)\mathbin{::}\Varid{f}\;(\Varid{a}\to \Varid{b})\to \Varid{f}\;\Varid{a}\to \Varid{f}\;\Varid{b}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%
+The \emph{Applicative} class describes how to compose effectul
+computations encoded as values of type $f\ a$ (the effectful
+computation of a pure value of type $a$). Thus, \emph{pure} constructs
+a trivially effectful computation from a pure value and \ensuremath{<\!\!\!*\!\!\!>} (akin to
+application) takes an effectful computation of a function and an
+effectful computation of an argument and evaluates the two effects.
+
+Our \emph{Temporal} denotations have the applicative functor
+structure with the following derivation in terms of the monad:
+%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{instance}\;\Conid{Functor}\;\Conid{Temporal}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{fmap}\;\Varid{f}\;\Varid{x}\mathrel{=}\mathbf{do}\;\{\mskip1.5mu \Varid{x'}\leftarrow \Varid{x};\Varid{return}\;(\Varid{f}\;\Varid{x'})\mskip1.5mu\}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\mathbf{instance}\;\Conid{Applicative}\;\Conid{Temporal}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{pure}\;\Varid{a}{}\<[21]%
+\>[21]{}\mathrel{=}\Varid{return}\;\Varid{a}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{f}<\!\!\!*\!\!\!>\Varid{x}{}\<[21]%
+\>[21]{}\mathrel{=}\mathbf{do}\;\{\mskip1.5mu \Varid{f'}\leftarrow \Varid{f};\Varid{x'}\leftarrow \Varid{x};\Varid{return}\;(\Varid{f'}\;\Varid{x'})\mskip1.5mu\}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%
+Note that the definition of \ensuremath{<\!\!\!*\!\!\!>} here orders the effects left-to-right.
+
+The interpretation of sequential composition for statements (with no
+dataflow) is then $\interp{P; Q} = (\lambda() \rightarrow
+\interp{P}) <\!\!\!*\!\!\!> \interp{Q}$.
+
+\paragraph{Monoid subset}
+
+Alternatively, the full structure of an applicative functor is
+not even needed in this restricted case. Instead, a monoid
+over \emph{Temporal ()} would suffice:
+%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{class}\;\Conid{Monoid}\;\Varid{m}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\Varid{mempty}\mathbin{::}\Varid{m}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\Varid{mappend}\mathbin{::}\Varid{m}\to \Varid{m}\to \Varid{m}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\mathbf{instance}\;\Conid{Monoid}\;(\Conid{Temporal}\;())\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{mempty}{}\<[19]%
+\>[19]{}\mathrel{=}\Varid{return}\;(){}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{a}\mathbin{`\Varid{mappend}`}\Varid{b}\mathrel{=}\mathbf{do}\;\{\mskip1.5mu \Varid{a};\Varid{b};\Varid{return}\;()\mskip1.5mu\}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%% 
+with the interpretation $\interp{P; Q} = \interp{P} \ensuremath{\mathbin{`\Varid{mappend}`}} Q$ and
+where \ensuremath{\Varid{mempty}} provides a \emph{no-op}.
 
@@
 
+\appendix
+
+\paragraph{Proof} (of Theorem~\ref{theorem:main})
+Since sequential composition is associative, we can reassociate
+$P; Q$ such that $P = P_1; P'$ where $P_1$ is a single statement
+
+\begin{itemize}
+\item $P = \sleep t$
+
+\begin{align*}
+\begin{array}{ll}
+       & \etime{\sleep t; Q} \\[0.5em]
+\equiv & \; \{\textit{Lemma}~\ref{lem:sleep-L}\} \\[0.1em]
+       & s + \etime{Q} \\[0.5em]
+\equiv & \; \{\textit{Definition}~\ref{}\} \\[0.1em]
+       & \etime{\sleep t} + \etime{Q}
+\end{array}
+\end{align*}
+
+\item $Q = \sleep t$ by Lemma~\ref{lem:sleep-R} then 
+there are two cases:
+
+\begin{itemize}
+\item $\etime{P} \leq t$ then:
+
+\begin{align*}
+\begin{array}{ll}
+       & \etime{P; \sleep t} \\[0.5em]
+\equiv & \; \{\textit{Lemma}~\ref{lem:sleep-R}\} \\[0.1em]
+       & \vtime{P} + t \\[0.5em]
+\equiv & \; \{\textit{Definition}~\ref{sleep-spec}\} \\[0.1em]
+       & \vtime{P} + \vtime{\sleep t}
+\end{array}
+\end{align*}
+
+\item $\etime{P} > t$ then:
+
+\begin{align*}
+\begin{array}{ll}
+       & \etime{P; \sleep t} \\[0.5em]
+\equiv & \; \{\textit{Lemma}~\ref{lem:sleep-R}\} \\[0.1em]
+       & \etime{P} \\[0.5em]
+\leq   & \etime{P} + \etime{\sleep t}
+\end{array}
+\end{align*}
+
+\end{itemize}
+
+\item $P = P';P'', Q = Q';Q''$
+
+Ressociate so that $P'; (P''; (Q'; Q''))$ then
+
+By induction:
+%%
+\begin{align}
+\vtime{P'} + \vtime{P''} \leq \etime{P'; P''} \leq \vtime{P'} + \vtime{P''} \\
+\vtime{Q'} + \vtime{Q''} \leq \etime{Q'; Q''} \leq \vtime{Q'} + \vtime{Q''} \\
+\end{align}
+
+and
+\note{STUCK!}
+\end{itemize}
+
 \end{document}

### 2014-04-29 time-monad.hs

@@ -1,39 +1,32 @@
-{-# LANGUAGE NoMonomorphismRestriction #-}
+{-# LANGUAGE NoMonomorphismRestriction, FlexibleInstances #-}
 
 import Data.Time
 import Control.Applicative
 import Control.Concurrent
+import Data.Monoid
 
 import Debug.Trace
 
+-- Some aliases to simplify code
 type Time = UTCTime
 type VTime = NominalDiffTime
 diffTime = diffUTCTime
 
+-- Core data type and monad definition
 data Temporal a = T ((Time, Time) -> (VTime -> IO (a, VTime)))
 
-instance Functor Temporal where
-    fmap f (T x) = T (\(startT, nowT) -> \vT -> do (a, vT') <- x (startT, nowT) vT
-                                                   return (f a, vT'))
-
-instance Applicative Temporal where
-    pure a          = T (\(_, _) -> \vt -> return (a, vt))
-    (T f) <*> (T x) = T (\(startT, nowT) -> \vT -> 
-                             do thenT      <- getCurrentTime
-                                (f', vT')  <- f (startT, thenT) vT
-                                (x', vT'') <- x (startT, nowT) vT'
-                                return (f' x', vT''))
-
 instance Monad Temporal where
-    return a     = T $ \(_, _) -> \vt -> return (a, vt)
-    (T c1) >>= f = T $ \(startT, nowT) -> \vT -> do thenT <- getCurrentTime
-                                                    (x, vT') <- c1 (startT, thenT) vT
-                                                    let (T c2) = f x in c2 (startT, nowT) vT'
-                                           
+    return a     = T (\_ -> \vt -> return (a, vt))
+
+    (T p) >>= q  = T (\(startT, nowT) -> \vt -> 
+                             do (x, vt')    <- p (startT, nowT) vt
+                                let (T q')  = q x
+                                thenT       <- getCurrentTime
+                                q' (startT, thenT) vt')
 
 runTime :: Temporal a -> IO a
-runTime (T c) = do t <- getCurrentTime
-                   (y, _) <- c (t, t) 0
+runTime (T c) = do startT <- getCurrentTime
+                   (y, _) <- c (startT, startT) 0
[…]

### 2014-04-30 paper/paper.tex
moar

@@
 \usepackage{amsthm}
+\usepackage{amssymb}
 \usepackage{amsmath}
@@
 \newtheorem{theorem}{Theorem}
+
+\makeatletter
+\newtheorem*{rep@lemma}{\rep@title}
+\newcommand{\newreplemma}[2]{%
+\newenvironment{rep#1}[1]{%
+ \def\rep@title{#2 \ref{##1} (recap)}%
+ \begin{rep@lemma}}%
+ {\end{rep@lemma}}}
+\makeatother
+
+\newreplemma{lemma}{Lemma}
+
 \theoremstyle{definition}
 \newtheorem{definition}{Definition}
+
+\newreplemma{definition}{Definition}
 
@@
 The programming model of \lang{} distinguishes between the
-\emph{actual time} elapsed since the start of a program $P$, which we
-write as $\etime{P}$, and the \emph{virtual time} which is advanced by
-\sleepOp{} statements, written $\vtime{P}$.
+\emph{actual time} elapsed since the start of a program $P$ which we
+write as $\etime{P}$ and the \emph{virtual time} which is advanced by
+\sleepOp{} statements which we write as $\vtime{P}$.
 
 Since \sleepOp{} is the only operation that changes the virtual
-time, we can specification for $\vtime{-}$ over all programs:
+time, we can easily specify the definition of $\vtime{-}$ over all programs:
 %
@@
 
-Equality of computation times is difficult. In order to not sweep this
-issue entirely under the carpet, we will use the relation $\approx$,
-where $s \approx t$ means that $s$ and $t$ are either exactly equal or
-differ by some amount $\epsilon$.\note{Discuss this further, may be
+Providing exact deadlines in real-time system is difficult due 
+to noise and overhead caused by execution. We do not wish to sweep
+this under the carpet in \lang{} or in our models. We therefore define
+a relation $\approx$ on actual time (non virtual times), where:
+%%
+\begin{equation}
+s \approx t
+\equiv
+t - \varepsilon \leq s \leq t + \varepsilon 
+\end{equation}
+%
+for some value of $\varepsilon$ which is the maximum negligible 
+time value with respect to the application at hand. In the case of
+\lang{}, this is equal to the scheduling delay for the SuperCollider
+engine (and is machine dependent), which is roughly \note{X}.
+
+\note{Discuss this further, may be
   able to say later that in some cases $\epsilon$ is the scheduling
-  time for play statments?}  The virtual time and actual time of a
-single sleep statement (in isolation) are roughly the same, \ie{},
-$\vtime{\sleep t} \approx \etime{\sleep t}$ and thus $\vtime{\sleep t}
-\approx t$ by the specification in Definition~\ref{sleep-spec}.
-
-We should be clear however, that this approximate equality does not always
-hold. We can only say that it holds when \sleepOp{} is used in isolation, that is, 
-when it is the only statement in a program. As shown by the examples
-of Section~\ref{sec:examples}, the use of $\sleep t$ in a program 
-does not mean that a program necessarily waits for $t$ seconds-- 
-depending on the context, it may wait for anywhere between $0$ and $t$ seconds.
-
-For convenience, and to contrast with \sleepOp{}, we'll use an additional
-statement \ksleepOp{} here (which is not available in the actual language)
- which always sleeps for the number of seconds specified by its parameter.
+  time for play statments?}
+
+The virtual time and actual time of a single sleep statement (in
+isolation) are roughly the same, \ie{}, $\vtime{\sleep t} \approx
+\etime{\sleep t}$ and thus $\etime{\sleep t} \approx t$ (by the
+specification in Definition~\ref{sleep-spec}). Note that this holds
+only when \sleepOp{} is used in isolation, that is, when it is the
+only statement in a program. As shown by the examples of
+Section~\ref{sec:examples}, the use of $\sleep t$ in a program does
+not mean that a program necessarily waits for $t$ seconds-- depending
+on the context, it may wait for anywhere between $0$ and $t$ seconds.
+
+%For convenience, and to contrast with \sleepOp{}, we'll use an additional
+%statement \ksleepOp{} here (which is not available in the actual language)
+% which always sleeps for the number of seconds specified by its parameter.
 
@@
 \begin{align*}
-\etime{P; \sleep{} t} = 
+\etime{P; \sleep{} t} \approx
  \begin{cases}
@@
 \begin{align*}
-\etime{\sleep{} t; P} = t + \etime{P}
+\etime{\sleep{} t; P} \approx t + \etime{P}
 \end{align*}
@@
 For all programs $P$ then $\etime{P} \geq \vtime{P}$. 
-\label{lemma1}
+\label{lemma-rel-etime-vtime}
 \end{lemma}
@@
 \begin{equation}
-\vtime{P} + \vtime{Q} \leq \etime{P; Q} \leq \etime{P} + \etime{Q}
+\vtime{P} + \vtime{Q} \leq \etime{P; Q} \lesssim \etime{P} + \etime{Q}
 \end{equation}
@@
 %
+To ease understanding, we recall the types of \emph{return}
+and \ensuremath{(\bind )} along with some intuition for their behaviour for
+\emph{Temporal}:
+%
 \begin{itemize}
-
-\item \ensuremath{\Varid{return}\mathbin{::}\Varid{a}\to \Conid{Temporal}\;\Varid{a}} thus ignores the time parameters and
-has the usual pure state behaviour: passing the state unchanged.
-For the bind
-
-\item \ensuremath{(\bind )\mathbin{::}\Conid{Temporal}\;\Varid{a}\to (\Varid{a}\to \Conid{Temporal}\;\Varid{b})\to \Conid{Temporal}\;\Varid{b}} thus
-composes two computations together. 
-The result of composing two temporal computations, with
-start time \emph{starT}, current time \emph{nowT}, and virtual
-time \emph{vT}, is the result of evaluating first the right 
-hand side, and then the left hand side
-
+\item \ensuremath{\Varid{return}\mathbin{::}\Varid{a}\to \Conid{Temporal}\;\Varid{a}} lifts a pure value into a trivially
+effectful computation by ignoring the time parameters and 
+providing the usual pure state behaviour: passing the state unchanged.
+
+\item \ensuremath{(\bind )\mathbin{::}\Conid{Temporal}\;\Varid{a}\to (\Varid{a}\to \Conid{Temporal}\;\Varid{b})\to \Conid{Temporal}\;\Varid{b}} 
+  composes two computations together.  The result of composing two
+  temporal computations, with start time \emph{startT}, current time
+  \emph{nowT}, and virtual time \emph{vT}, is the result of evaluating
+  first the left-hand side at time \emph{nowT} and then
+
+right hand side, and then the left hand side.  Note that
+  \ensuremath{\Varid{getCurrentTime}\mathbin{::}\Conid{IO}\;\Conid{Time}} retrieves the time from the operation
+  system.
 \end{itemize}
 
-The following \ensuremath{\Varid{runTime}} operation executes a temporal computation
+The \ensuremath{\Varid{runTime}} operation executes a temporal computation
 inside of the \emph{IO} monad by providing the start time of the
@@
 (where \ensuremath{\Varid{f}\mathrel{=}\Conid{T}\;\Varid{f'},\Varid{g}\mathrel{=}\Conid{T}\;\Varid{g'},\Varid{h}\mathrel{=}\Conid{T}\;\Varid{h'}}) which 
-desugars to the following, after some simplification: 
+desugars to the following \emph{IO} computation, 
+after some simplification: 
 %%
@@
 throughout the computation.
+
+Figure~\ref{core-functions} provides a number of effectul operations
+for \emph{Temporal} that access the current time, the start time, get
+and set the virtual time, and cause a kernel sleep. These
+are used in the next part of the model.
 
@@
 \caption{Simple \emph{Temporal} computations, used  by the model}
+\label{core-functions}
 \end{figure}
@@
 \column{20}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{28}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{30}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{37}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[B]{}\Varid{sleep}\;\Varid{delayT}\mathrel{=}\mathbf{do}\;{}\<[20]%
-\>[20]{}\Varid{vT}{}\<[28]%
-\>[28]{}\leftarrow \Varid{getVirtualTime}{}\<[E]%
-\\
-\>[20]{}\Varid{startT}{}\<[28]%
-\>[28]{}\leftarrow \Varid{start}{}\<[E]%
-\\
-\>[20]{}\Varid{nowT}{}\<[28]%
-\>[28]{}\leftarrow \Varid{time}{}\<[E]%
+\>[20]{}\Varid{nowT}{}\<[30]%
+\>[30]{}\leftarrow \Varid{time}{}\<[E]%
+\\
+\>[20]{}\Varid{vT}{}\<[30]%
+\>[30]{}\leftarrow \Varid{getVirtualTime}{}\<[E]%
 \\
@@
 \\
+\>[20]{}\Varid{setVirtualTime}\;\Varid{vT'}{}\<[E]%
+\\
+\>[20]{}\Varid{startT}{}\<[30]%
+\>[30]{}\leftarrow \Varid{start}{}\<[E]%
+\\
 \>[20]{}\mathbf{let}\;\Varid{diffT}\mathrel{=}\Varid{diffTime}\;\Varid{nowT}\;\Varid{startT}{}\<[E]%
 \\
-\>[20]{}\mathbf{if}\;(\Varid{vT'}\mathbin{<}\Varid{diffT}){}\<[E]%
-\\
-\>[20]{}\hsindent{2}{}\<[22]%
-\>[22]{}\mathbf{then}\;\Varid{return}\;(){}\<[E]%
-\\
-\>[20]{}\hsindent{2}{}\<[22]%
-\>[22]{}\mathbf{else}\;\Varid{kernelSleep}\;(\Varid{vT'}\mathbin{-}\Varid{diffT}){}\<[E]%
-\\
-\>[20]{}\Varid{setVirtualTime}\;\Varid{vT'}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-%%
-Thus, \emph{sleep} proceeds by calculating the elapsed time, \emph{diffT}, 
-and the new virtual time \emph{vT'} (after the sleep). 
-If the new virtual time is less than the elapsed time then no actual (kernel) 
-sleeping happens. However, if the new virtual time is ahead of the elapsed time, 
-then the process sleeps for the difference. Finally, the virtual time state is updated
-with the new virtual time.
+\>[20]{}\mathbf{if}\;(\Varid{vT'}\mathbin{<}\Varid{diffT})\;\mathbf{then}\;\Varid{return}\;(){}\<[E]%
+\\
+\>[20]{}\hsindent{17}{}\<[37]%
+\>[37]{}\mathbf{else}\;\Varid{kernelSleep}\;(\Varid{vT'}\mathbin{-}\Varid{diffT}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%
+Thus, \emph{sleep} proceeds by getting the current time, calculating
+the new virtual time \emph{vT'} and updating the virtual time
+state. If the new virtual time is less than the elapsed time
+\emph{diffT} then no actual (kernel) sleeping happens. However, if the
+new virtual time is ahead of the elapsed time, then the process sleeps
+for the difference.
 
@@
 
-\paragraph{Lemma~\ref{lem:sleep-R}}
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
+We replay the previous lemmas on the temporal behaviour of \lang{} programs,
+and show that the monadic model is sound with respect to these, \ie{},
+that the lemmas hold of the model.
+
+\noindent
+\begin{repdefinition}{sleep-spec}
+\begin{align*}
+\vtime{\sleep t} & = t \\ 
+\vtime{P; Q} & = \vtime{P} + \vtime{Q} \\
+\vtime{-} & = 0
+\end{align*}
+\end{repdefinition}
+
+\begin{proof}
+For our model, the proof is straightforward. For the case of
+$P; Q$, we rely on the monotonicity of virtual time: virtual
+time is only ever increasing, and can only ever be incremented by sleep. 
+\note{Could put more here}
+\end{proof}
+
+\begin{replemma}{lem:sleep-R}
+For some program $P$ and time $t$: 
+%%
+\begin{align*}
+\etime{P; \sleep{} t} \approx
+ \begin{cases}
+   \etime{P} & (\vtime{P} + t) < \etime{P} \\
+   \vtime{P} + t  & \textit{otherwise}
+ \end{cases}
+\end{align*}
+\end{replemma}
+
+\begin{proof}
+Our model interprets $(P; \sleep t)$ as:
+%%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{runTime}\;(\mathbf{do}\;\{\mskip1.5mu \interp{P};\Varid{sleep}\;\Varid{t}\mskip1.5mu\}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%%
+which desugars and simplifies to the following \emph{IO} computation:
+%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{7}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{16}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{do}\;\Varid{startT}\leftarrow \Varid{getCurrentTime}{}\<[E]%
-\\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}(\Varid{x},\Varid{vT'}){}\<[16]%
-\>[16]{}\leftarrow \Varid{p}\;(\Varid{startT},\Varid{startT})\;\mathrm{0}{}\<[E]%
-\\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}\Varid{thenT}{}\<[16]%
+\>[B]{}\mathbf{do}\;{}\<[5]%
+\>[5]{}\Varid{startT}{}\<[16]%
 \>[16]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
 \\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}\mathbf{let}\;\Varid{vT''}{}\<[15]%
-\>[15]{}\mathrel{=}\Varid{vT'}\mathbin{+}\Varid{t}{}\<[E]%
-\\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}\mathbf{let}\;\Varid{diffT}\mathrel{=}\Varid{diffTime}\;\Varid{thenT}\;\Varid{startT}{}\<[E]%
-\\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}\mathbf{if}\;(\Varid{vT''}\mathbin{<}\Varid{diffT}){}\<[E]%
-\\
-\>[4]{}\hsindent{3}{}\<[7]%
+\>[5]{}(\Varid{x},\Varid{vT'}){}\<[16]%
+\>[16]{}\leftarrow \interp{P}\;(\Varid{startT},\Varid{startT})\;\mathrm{0}{}\<[E]%
+\\
+\>[5]{}\Varid{nowT}{}\<[16]%
+\>[16]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
+\\
+\>[5]{}\mathbf{let}\;\Varid{vT''}{}\<[16]%
+\>[16]{}\mathrel{=}\Varid{vT'}\mathbin{+}\Varid{t}{}\<[E]%
+\\
+\>[5]{}\Varid{setVirtualTime}\;\Varid{vT''}{}\<[E]%
+\\
+\>[5]{}\mathbf{let}\;\Varid{diffT}{}\<[16]%
+\>[16]{}\mathrel{=}\Varid{diffTime}\;\Varid{nowT}\;\Varid{startT}{}\<[E]%
+\\
+\>[5]{}\mathbf{if}\;(\Varid{vT''}\mathbin{<}\Varid{diffT}){}\<[E]%
+\\
+\>[5]{}\hsindent{2}{}\<[7]%
 \>[7]{}\mathbf{then}\;\Varid{return}\;(){}\<[E]%
 \\
-\>[4]{}\hsindent{3}{}\<[7]%
-\>[7]{}\mathbf{else}\;\Varid{kernelSleep}\;(\Varid{vT''}\mathbin{-}\Varid{diffT}){}\<[E]%
-\\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}\Varid{setVirtualTime}\;\Varid{vT''}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
+\>[5]{}\hsindent{2}{}\<[7]%
+\>[7]{}\mathbf{else}\;\Varid{kernelSleep'}\;(\Varid{vT''}\mathbin{-}\Varid{diffT}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%%
+where \ensuremath{\Varid{kernelSleep'}\;\Varid{x}\mathrel{=}\Varid{threadDelay}\;(\Varid{round}\;(\Varid{x}\mathbin{*}\mathrm{1000000}))} is used
+as an alias here (as per the definition of \emph{kernelSleep} in
+Figure~\ref{core-functions}.
+
+From this we see that $\emph{diffT} = \etime{P}$ and $\emph{vT'} = \vtime{P}$ and 
+ $\emph{vT''} = \vtime{P} + t$. Therefore, the guard of the 
+\emph{if} expression is $(\vtime{P} + t) < \etime{P}$.
+If the updating of the virtual time state and the computing of 
+the guard takes $e$ then the overall time taken is:
+%%
 \begin{align*}
@@
  \begin{cases}
-   \etime{P} & (\vtime{P} + t) < \etime{P}  \\
-   \etime{P} + (\vtime{P} + t) - \etime{P}  & \textit{otherwise}
+   \etime{P} + e & (\vtime{P} + t) < \etime{P}  \\
+   \etime{P} + e + (\vtime{P} + t) - \etime{P}  & \textit{otherwise}
  \end{cases}
 \end{align*}
-
+%%
+which is equivalent to the statment of the lemma if $e \leq \epsilon$
+and if the reduction to the interpretation to get to the above code 
+takes less than $\epsilon$:
+\begin{align*}
+\etime{P; \sleep{} t} \approx 
+ \begin{cases}
+   \etime{P} & (\vtime{P} + t) < \etime{P} \\
+   \vtime{P} + t  &  \textit{otherwise}
+ \end{cases}
+\end{align*}
+\end{proof}
+
+\begin{replemma}{lem:sleep-L}
+For some program $P$ and time $t$:
+%%
+\begin{align*}
+\etime{\sleep{} t; P} \approx t + \etime{P}
+\end{align*}
+\end{replemma}
+
+\begin{proof}
+Our models interprets $(\sleep t; P)$ as:
+%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{runTime}\;(\mathbf{do}\;\{\mskip1.5mu \Varid{sleep}\;\Varid{t};\interp{P}\mskip1.5mu\}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%
+which desugars and simplifies to the following \emph{IO} computation:
+%
+%\begin{code}
+%do  setVirtualTime delayT
+%    kernelSleep delayT 
+%    interpP
+%\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{do}\;{}\<[5]%
+\>[5]{}\Varid{startT}\leftarrow \Varid{getCurrentTime}{}\<[E]%
+\\
+\>[5]{}\Varid{kernelSleep}\;\Varid{delayT}{}\<[E]%
+\\
+\>[5]{}\mathbf{let}\;(\Conid{T}\;\Varid{p})\mathrel{=}\interp{P}\;(){}\<[E]%
+\\
+\>[5]{}\Varid{thenT}\leftarrow \Varid{getCurrentTime}{}\<[E]%
+\\
+\>[5]{}\Varid{p}\;(\Varid{startT},\Varid{thenT})\;\Varid{delayT}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%
+The code does a \ensuremath{\Varid{kernelSleep}} for $t$ and then continues with $P$ at
+virtual time $t$, \ie{}, the time taken is $t + \interp{P}_t$.  In
+desugaring and simplifying we have certainly elided some intermediate
+steps in the computation (such as the guard test in sleep), which we
+account for as part of the error $\epsilon$.
+\end{proof}
+
+\begin{replemma}{lemma-rel-etime-vtime}
+For all programs $P$ then $\etime{P} \geq \vtime{P}$. 
+\end{replemma}
+
+\begin{proof}
+By induction % possibly by strong induction?
+
+$\interp{sleep t} = $
+\ensuremath{\Varid{runTime}\;(\Varid{sleep}\;\Varid{t})} $\leadsto$ \ensuremath{\Varid{kernelSleep'}\;\Varid{delayT}}
+
+$\etime{P;Q} = $
+Inductive hypothesis: $\etime{P} \geq \vtime{P}$
+              $\etime{Q} \geq \vtime{Q}$
+
+\begin{itemize}
+\item case $P = sleep t$
+
+$\etime{sleep t; Q} = t + \etime{Q}$ 
+$\vtime{sleep t; Q} = t + \vtime{Q}$
+by inductive hypothesis $t + \etime{Q} \geq t + \vtime{Q}$.
+
+\item case $Q = sleep t$
+
+$\vtime{P; sleep t} = \vtime{P} + t$
+By inductive hypothesis ($\etime{P} \geq \vtime{P}$) then
+
+\begin{itemize}
+\item case $\vtime{P} + t \leq \etime{P}$ then
+
+$\etime{P; sleep t} = \etime{P}$ 
+therefore $\etime{P; sleep t} \geq \vtime{P; sleep t}$
+since $\etime{P} \geq \vtime{P} + t$ by the case.
+
+\item case $\vtime{P} + t \geq \etime{P}$ then
+ $\etime{P; sleep t} = \vtime{P} + t$
+then $\etime{P; sleep t} \geq \vtime{P; sleep t}$
+since $\etime{P; sleep t} = \vtime{P; sleep t}$. 
+\end{itemize}
+
+\item case $P = P';P''$, $Q = Q';Q''$
+
+reassociate
+
+case P' = sleep t,
+
+$\etime{P';Q'} = t + \etime{Q'}$
+\end{itemize}
+
+\end{proof}
+
+\subsubsection{Monad laws}
+
+\paragraph{Quotienting by non-time dependent functions}
+
+\note{This section is more of a marker for myself (Dom), I need
+to think about this later, but basically it is about showing
+that the monad laws hold for our semantics (but not in general)}
+
+In L, there is no expression which returns the current time; 
+ \emph{getTime} belongs only to the model, not to the language.
+That is, for all expressions $e$, then the denotation 
+$\interp{e}$ factors through
+
+\subsection{Subsets of the semantics}
+
+For the examples of Section~\ref{sec:introduction}, the full structure
+of monad is not needed to give their semantics as there is no using of
+binding between statements (and thus no dataflow). In these case just
+an \emph{applicative functor}~\cite{mcbride2008functional} or even a
+monoid would suffice. These can be derived from the monad structure
+on \emph{Temporal} since all monads are applicative functors and all 
+monads define monoid over $m ()$.
+
+\paragraph{Applicative subset}
+
+Applicative functors are described by the following interface in
+Haskell:
+%%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{class}\;\Conid{Functor}\;\Varid{f}\Rightarrow \Conid{Applicative}\;\Varid{f}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\Varid{pure}{}\<[10]%
+\>[10]{}\mathbin{::}\Varid{a}\to \Varid{f}\;\Varid{a}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}(<\!\!\!*\!\!\!>)\mathbin{::}\Varid{f}\;(\Varid{a}\to \Varid{b})\to \Varid{f}\;\Varid{a}\to \Varid{f}\;\Varid{b}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%
+The \emph{Applicative} class describes how to compose effectul
+computations encoded as values of type $f\ a$ (the effectful
+computation of a pure value of type $a$). Thus, \emph{pure} constructs
+a trivially effectful computation from a pure value and \ensuremath{<\!\!\!*\!\!\!>} (akin to
+application) takes an effectful computation of a function and an
+effectful computation of an argument and evaluates the two effects.
+
+Our \emph{Temporal} denotations have the applicative functor
+structure with the following derivation in terms of the monad:
+%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{instance}\;\Conid{Functor}\;\Conid{Temporal}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{fmap}\;\Varid{f}\;\Varid{x}\mathrel{=}\mathbf{do}\;\{\mskip1.5mu \Varid{x'}\leftarrow \Varid{x};\Varid{return}\;(\Varid{f}\;\Varid{x'})\mskip1.5mu\}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\mathbf{instance}\;\Conid{Applicative}\;\Conid{Temporal}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{pure}\;\Varid{a}{}\<[21]%
+\>[21]{}\mathrel{=}\Varid{return}\;\Varid{a}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{f}<\!\!\!*\!\!\!>\Varid{x}{}\<[21]%
+\>[21]{}\mathrel{=}\mathbf{do}\;\{\mskip1.5mu \Varid{f'}\leftarrow \Varid{f};\Varid{x'}\leftarrow \Varid{x};\Varid{return}\;(\Varid{f'}\;\Varid{x'})\mskip1.5mu\}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%
+Note that the definition of \ensuremath{<\!\!\!*\!\!\!>} here orders the effects left-to-right.
+
+The interpretation of sequential composition for statements (with no
+dataflow) is then $\interp{P; Q} = (\lambda() \rightarrow
+\interp{P}) <\!\!\!*\!\!\!> \interp{Q}$.
+
+\paragraph{Monoid subset}
+
+Alternatively, the full structure of an applicative functor is
+not even needed in this restricted case. Instead, a monoid
+over \emph{Temporal ()} would suffice:
+%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{class}\;\Conid{Monoid}\;\Varid{m}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\Varid{mempty}\mathbin{::}\Varid{m}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\Varid{mappend}\mathbin{::}\Varid{m}\to \Varid{m}\to \Varid{m}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\mathbf{instance}\;\Conid{Monoid}\;(\Conid{Temporal}\;())\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{mempty}{}\<[19]%
+\>[19]{}\mathrel{=}\Varid{return}\;(){}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{a}\mathbin{`\Varid{mappend}`}\Varid{b}\mathrel{=}\mathbf{do}\;\{\mskip1.5mu \Varid{a};\Varid{b};\Varid{return}\;()\mskip1.5mu\}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%% 
+with the interpretation $\interp{P; Q} = \interp{P} \ensuremath{\mathbin{`\Varid{mappend}`}} Q$ and
+where \ensuremath{\Varid{mempty}} provides a \emph{no-op}.
+
+\subsection{Alternate definition of \emph{sleep}}
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{31}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{sleep'}\mathbin{::}\Conid{VTime}\to \Conid{Temporal}\;(){}\<[E]%
+\\
+\>[B]{}\Varid{sleep'}\;\Varid{delayT}\mathrel{=}\mathbf{do}\;{}\<[21]%
+\>[21]{}\Varid{vT}{}\<[31]%
+\>[31]{}\leftarrow \Varid{getVirtualTime}{}\<[E]%
+\\
+\>[21]{}\mathbf{let}\;\Varid{vT'}{}\<[31]%
+\>[31]{}\mathrel{=}\Varid{vT}\mathbin{+}\Varid{delayT}{}\<[E]%
+\\
+\>[21]{}\Varid{setVirtualTime}\;\Varid{vT'}{}\<[E]%
+\\
+\>[21]{}\Varid{startT}{}\<[31]%
+\>[31]{}\leftarrow \Varid{start}{}\<[E]%
+\\
+\>[21]{}\Varid{nowT}{}\<[31]%
+\>[31]{}\leftarrow \Varid{time}{}\<[E]%
+\\
+\>[21]{}\mathbf{let}\;\Varid{diffT}\mathrel{=}\Varid{diffTime}\;\Varid{nowT}\;\Varid{startT}{}\<[E]%
+\\
+\>[21]{}\mathbf{if}\;(\Varid{vT'}\mathbin{<}\Varid{diffT}){}\<[E]%
+\\
+\>[21]{}\hsindent{4}{}\<[25]%
+\>[25]{}\mathbf{then}\;\Varid{return}\;(){}\<[E]%
+\\
+\>[21]{}\hsindent{4}{}\<[25]%
+\>[25]{}\mathbf{else}\;\Varid{kernelSleep}\;(\Varid{vT'}\mathbin{-}\Varid{diffT}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%
+This alternate definition should reduce any oversleeping by minimising 
+noise in the timing. For example, the virtual time is calculated 
+and updated before the current time is retrieved in case the additional
+time taken in updating the virtual time means that the elapsed time
+catches up with the virtual time.
+
+To see the difference, consider Lemma~\ref{lem:sleep-R}:
+
+\begin{replemma}{lem:sleep-R}
+For some program $P$ and time $t$: 
+%%
 \begin{align*}
@@
 \end{align*}
-
-\paragraph{Lemma~\ref{lem:sleep-L}}
- $\etime{\sleep{} t; P} = t + \etime{P}$.
-This lemma holds in the model, where \ensuremath{\Varid{runTime}\;(\mathbf{do}\;\{\mskip1.5mu \Varid{sleep}\;\Varid{t};\Conid{P}\mskip1.5mu\})} desugars and
-simplifies to:
+\end{replemma}
+%
+\noindent
+If the above alternate definition \emph{sleep'} is used, then
+the interpreation of $(P; \sleep t)$  
+desugars and simplifies to the following:
 %
@@
 \column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
 \>[B]{}\mathbf{do}\;{}\<[5]%
-\>[5]{}\Varid{startT}\leftarrow \Varid{getCurrentTime}{}\<[E]%
-\\
-\>[5]{}\Varid{kernelSleep}\;\Varid{t}{}\<[E]%
-\\
-\>[5]{}\mathbf{let}\;(\Conid{T}\;\Varid{p'}){}\<[17]%
-\>[17]{}\mathrel{=}\Varid{p}\;(){}\<[E]%
-\\
-\>[5]{}\Varid{thenT}{}\<[12]%
-\>[12]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
-\\
-\>[5]{}\Varid{p'}\;(\Varid{startT},\Varid{thenT})\;\Varid{t}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-%
-Thus, we see that we \ensuremath{\Varid{kernelSleep}} for $t$, and then continue with $P$ at virtual
-time $t$.
-
-\paragraph{Quotienting by non-time dependent functions}
-
-\note{This section is more of a marker for myself (Dom), I need
-to think about this later, but basically it is about showing
-that the monad laws hold for our semantics (but not in general)}
-
-In L, there is no expression which returns the current time; 
- \emph{getTime} belongs only to the model, not to the language.
-That is, for all expressions $e$, then the denotation 
-$\interp{e}$ factors through
-
-\subsection{Subsets of the semantics}
-
-For the examples of Section~\ref{sec:introduction}, the full structure
-of monad is not needed to give their semantics as there is no using of
-binding between statements (and thus no dataflow). In these case just
-an \emph{applicative functor}~\cite{mcbride2008functional} or even a
-monoid would suffice. These can be derived from the monad structure
-on \emph{Temporal} since all monads are applicative functors and all 
-monads define monoid over $m ()$.
-
-\paragraph{Applicative subset}
-
-Applicative functors are described by the following interface in
-Haskell:
-%%
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{class}\;\Conid{Functor}\;\Varid{f}\Rightarrow \Conid{Applicative}\;\Varid{f}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}\Varid{pure}{}\<[10]%
-\>[10]{}\mathbin{::}\Varid{a}\to \Varid{f}\;\Varid{a}{}\<[E]%
-\\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}(<\!\!\!*\!\!\!>)\mathbin{::}\Varid{f}\;(\Varid{a}\to \Varid{b})\to \Varid{f}\;\Varid{a}\to \Varid{f}\;\Varid{b}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-%
-The \emph{Applicative} class describes how to compose effectul
-computations encoded as values of type $f\ a$ (the effectful
-computation of a pure value of type $a$). Thus, \emph{pure} constructs
-a trivially effectful computation from a pure value and \ensuremath{<\!\!\!*\!\!\!>} (akin to
-application) takes an effectful computation of a function and an
-effectful computation of an argument and evaluates the two effects.
-
-Our \emph{Temporal} denotations have the applicative functor
-structure with the following derivation in terms of the monad:
-%
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{instance}\;\Conid{Functor}\;\Conid{Temporal}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{fmap}\;\Varid{f}\;\Varid{x}\mathrel{=}\mathbf{do}\;\{\mskip1.5mu \Varid{x'}\leftarrow \Varid{x};\Varid{return}\;(\Varid{f}\;\Varid{x'})\mskip1.5mu\}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\mathbf{instance}\;\Conid{Applicative}\;\Conid{Temporal}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{pure}\;\Varid{a}{}\<[21]%
-\>[21]{}\mathrel{=}\Varid{return}\;\Varid{a}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{f}<\!\!\!*\!\!\!>\Varid{x}{}\<[21]%
-\>[21]{}\mathrel{=}\mathbf{do}\;\{\mskip1.5mu \Varid{f'}\leftarrow \Varid{f};\Varid{x'}\leftarrow \Varid{x};\Varid{return}\;(\Varid{f'}\;\Varid{x'})\mskip1.5mu\}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-%
-Note that the definition of \ensuremath{<\!\!\!*\!\!\!>} here orders the effects left-to-right.
-
-The interpretation of sequential composition for statements (with no
-dataflow) is then $\interp{P; Q} = (\lambda() \rightarrow
-\interp{P}) <\!\!\!*\!\!\!> \interp{Q}$.
-
-\paragraph{Monoid subset}
-
-Alternatively, the full structure of an applicative functor is
-not even needed in this restricted case. Instead, a monoid
-over \emph{Temporal ()} would suffice:
-%
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{class}\;\Conid{Monoid}\;\Varid{m}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}\Varid{mempty}\mathbin{::}\Varid{m}{}\<[E]%
-\\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}\Varid{mappend}\mathbin{::}\Varid{m}\to \Varid{m}\to \Varid{m}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\mathbf{instance}\;\Conid{Monoid}\;(\Conid{Temporal}\;())\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{mempty}{}\<[19]%
-\>[19]{}\mathrel{=}\Varid{return}\;(){}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{a}\mathbin{`\Varid{mappend}`}\Varid{b}\mathrel{=}\mathbf{do}\;\{\mskip1.5mu \Varid{a};\Varid{b};\Varid{return}\;()\mskip1.5mu\}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-%% 
-with the interpretation $\interp{P; Q} = \interp{P} \ensuremath{\mathbin{`\Varid{mappend}`}} Q$ and
-where \ensuremath{\Varid{mempty}} provides a \emph{no-op}.
+\>[5]{}\Varid{startT}{}\<[16]%
+\>[16]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
+\\
+\>[5]{}(\Varid{x},\Varid{vT'}){}\<[16]%
+\>[16]{}\leftarrow \interp{P}\;(\Varid{startT},\Varid{startT})\;\mathrm{0}{}\<[E]%
+\\
+\>[5]{}\mathbf{let}\;\Varid{vT''}{}\<[16]%
+\>[16]{}\mathrel{=}\Varid{vT'}\mathbin{+}\Varid{t}{}\<[E]%
+\\
+\>[5]{}\Varid{setVirtualTime}\;\Varid{vT''}{}\<[E]%
+\\
+\>[5]{}\Varid{nowT}{}\<[16]%
+\>[16]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
+\\
+\>[5]{}\mathbf{let}\;\Varid{diffT}{}\<[16]%
+\>[16]{}\mathrel{=}\Varid{diffTime}\;\Varid{nowT}\;\Varid{startT}{}\<[E]%
+\\
+\>[5]{}\mathbf{if}\;(\Varid{vT''}\mathbin{<}\Varid{diffT}){}\<[E]%
+\\
+\>[5]{}\hsindent{2}{}\<[7]%
+\>[7]{}\mathbf{then}\;\Varid{return}\;(){}\<[E]%
+\\
+\>[5]{}\hsindent{2}{}\<[7]%
+\>[7]{}\mathbf{else}\;\Varid{kernelSleep}\;(\Varid{vT''}\mathbin{-}\Varid{diffT}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%%
+which exhibits the temporal behaviour:
+%% 
+\begin{align*}
+\etime{P; \sleep{} t} = 
+ \begin{cases}
+   \etime{P} + e_1 + e_2 & (\vtime{P} + t) < (\etime{P} + e_1) \\
+   \vtime{P} + t + e_2 &  \textit{otherwise}
+ \end{cases}
+\end{align*}
+%
+where $e_1$ is the time taken by updating the virtual time and
+$e_2$ is the time taken by the guard. This gives
+a tighter bound on sleep behaviour that previously where the behaviour was:
+%
+\begin{align*}
+\etime{P; \sleep{} t} = 
+ \begin{cases}
+   \etime{P} + e_1 + e_2 & (\vtime{P} + t) < \etime{P} \\
+   \vtime{P} + t + e_1 + e_2 &  \textit{otherwise}
+ \end{cases}
+\end{align*}
+%
 

### 2014-04-30 Tests.hs

@@ -1,135 +1,4 @@
-{-# LANGUAGE NoMonomorphismRestriction, FlexibleInstances #-}
-
-import Data.Time
-import Control.Applicative
-import Control.Concurrent
-import Data.Monoid
-
-import Debug.Trace
-
--- Some aliases to simplify code
-type Time = UTCTime
-type VTime = NominalDiffTime
-diffTime = diffUTCTime
-
--- Core data type and monad definition
-data Temporal a = T ((Time, Time) -> (VTime -> IO (a, VTime)))
-
-instance Monad Temporal where
-    return a     = T (\_ -> \vt -> return (a, vt))
-
-    (T p) >>= q  = T (\(startT, nowT) -> \vt -> 
-                             do (x, vt')    <- p (startT, nowT) vt
-                                let (T q')  = q x
-                                thenT       <- getCurrentTime
-                                q' (startT, thenT) vt')
-
-runTime :: Temporal a -> IO a
-runTime (T c) = do startT <- getCurrentTime
-                   (y, _) <- c (startT, startT) 0
-                   return y
-
-time, start :: Temporal Time
-time  = T (\(_, nowT) -> \vT -> return (nowT, vT))
-start = T (\(startT, _) -> \vT -> return (startT, vT))
-
-getVirtualTime :: Temporal VTime
-getVirtualTime = T (\(_, _) -> \vT -> return (vT, vT))
-
-setVirtualTime :: VTime -> Temporal ()
-setVirtualTime vT = T (\_ -> \_ -> return ((), vT))
-
-
-sleep :: VTime -> Temporal ()
-sleep delayT = do nowT      <- time
-                  vT        <- getVirtualTime
-                  let vT'   = vT + delayT
-                  setVirtualTime vT'
-                  startT    <- start
-                  let diffT = diffTime nowT startT
-                  if (vT' < diffT) then return () 
-                                   else kernelSleep (vT' - diffT)
-
-
-{-
-sleept :: VTime -> Temporal ()
-sleept delayT = do vT        <- getVirtualTime
-                   startT    <- start
-                   nowT      <- time
-                   let vT'   = vT + delayT
-                   let diffT = diffTime nowT startT
-                   if (vT' < diffT) then return () 
-                                   else kernelSleep (vT' - diffT)
[…]

### 2014-05-01 paper/paper.tex
generated .tex file

@@
 where three chords are played in sequence, combining simple notions of 
-parallel, timed, and ordered effects. The first three statements play
-the notes of a C major chord in parallel. A \sleepOp{} statement then 
-provides a \emph{barrier}
-
-Thus ``$\sleep{} t$'' communicates that, after it has been evaluated, at least 
+parallel, timed, and ordered effects.
+
+The first three statements play the notes of a C major chord in
+parallel.  A \sleepOp{} statement then provides a ``temporal barrier''
+which blocks the computation from continuing until 1 second has
+elapsed since the \emph{start} of the program (not since the end of
+playing the notes). Once one second has elapsed, the next three
+statements are executed which plays an F major chord. The next
+\sleepOp{} means that the final chord is not played until 1.5 seconds
+has elapsed since the start of the
+program. Figure~\ref{three-chord-timing} illustrates the timing.
+
+Thus, ``$\sleep{} t$'' communicates that, after it has been evaluated, at least 
 $t$ seconds has elapsed since the last \sleepOp{}. This provides a minimum
@@
 
+\note{Contributions}
+
 \begin{itemize}
-\item A computational model of virtual and real time 
-      structured by an \emph{applicative functor} abstraction.
-
-\item We show that the applicative approach extends
-      to a monadic approach, which can be composed with additional
-      monads to capture other useful notions of effect in \lang{} programs,
-      such as random numbers (Section~\ref{}).
+\item Explain the core principles of the language related to timing
+(Section~\ref{}) and provide a monadic model this programming approach
+(Section~\ref{}) (embedded in Haskell).
+
+\item We show that the monadic approach can be composed with
+  additional monads to capture other useful notions of effect in
+  \lang{} programs, such as random numbers (Section~\ref{}).
 \end{itemize}
 
-Previously reported on the language~\cite{aaron2013sonic}
+%
+The \lang{} was previously discussed in~\cite{aaron2013sonic}
 
@@
 Figure~\ref{sleep-examples} shows four similar programs which each
-have different internal behaviours for \sleepOp, illustrating the
-semantics of \sleepOp{}. The first three take 3s to execute and the
+have different internal behaviours for \sleepOp, illustrating its semantics. 
+The first three take 3s to execute and the
 last takes 4s to execute, with the behaviours:
@@
 \item{3s -- performs a computation lasting 1s, ignores
-the first \sleepOp{} since its minimum deadline has been reached, 
+the first \sleepOp{} since its minimum duration has been reached, 
 and then sleeps for 2s (one sleep performed);}
@@
 the first \sleepOp{} is ignored, and the second \sleepOp{} waits
-for only 1s to reach its minimum deadline (half a sleep performed);}
+for only 1s to reach its minimum duration (half a sleep performed);}
 \item{4s -- performs a computation lasting 2s, thus 
@@
 
-Since \sleepOp{} is the only operation that changes the virtual
-time, we can easily specify the definition of $\vtime{-}$ over all programs:
+The \sleepOp{} operation is the way to change the virtual
+time. Therefore, the definition of $\vtime{-}$ can be easily defined
+over all programs:
 %
@@
 \end{definition}
-
-Providing exact deadlines in real-time system is difficult due 
-to noise and overhead caused by execution. We do not wish to sweep
-this under the carpet in \lang{} or in our models. We therefore define
-a relation $\approx$ on actual time (non virtual times), where:
+\note{I haven't included here expressions, for example, if a call
+to a function that did some sleeping happened, then this is not accounted
+for here. I could be easily incldued though.}
+
+\paragraph{Equality on time}
+
+Providing exact deadlines in real-time systems is difficult due 
+to noise and overhead caused by execution. We do not ignore
+this problem in the programming model of \lang{} and the discussion here. 
+We define the relation $\approx$ on actual time (non virtual times), where:
 %%
@@
 s \approx t
+\;
 \equiv
-t - \varepsilon \leq s \leq t + \varepsilon 
+\;
+(t - \epsilon) \leq s \leq (t + \epsilon)
 \end{equation}
 %
-for some value of $\varepsilon$ which is the maximum negligible 
+for some value of $\epsilon$ which is the maximum negligible 
 time value with respect to the application at hand. In the case of
@@
 
-The virtual time and actual time of a single sleep statement (in
-isolation) are roughly the same, \ie{}, $\vtime{\sleep t} \approx
+The virtual time and actual time of a single sleep statement 
+ are roughly the same, \ie{}, $\vtime{\sleep t} \approx
 \etime{\sleep t}$ and thus $\etime{\sleep t} \approx t$ (by the
-specification in Definition~\ref{sleep-spec}). Note that this holds
+specification in Definition~\ref{sleep-spec}). This holds
 only when \sleepOp{} is used in isolation, that is, when it is the
@@
 on the context, it may wait for anywhere between $0$ and $t$ seconds.
+
+\paragraph{Temporal properties of programs}
+
+We outline here some important temporal properties of our \lang{} programs
+that relates the virtual time and actual times. In Section~\ref{}, we
+replay these lemmas and prove the soundness: that these lemmas hold of our model.
 
@@
 
+\noindent
+These two lemmas illuminate something of the semantics of sleep,
+and its interaction with other statements in the language.
+
 \begin{lemma}
@@
 
-\begin{theorem}
+\begin{lemma}
 For all programs $P$ and $Q$ then:
@@
 \label{theorem:main}
-\end{theorem}
-
-From these lemmas we can reason about the evaluation time of programs.
-For example, consider subprograms $A$, $B$, $C$ where 
-$\etime{A} = t_1$, $\etime{B} = t_2$, $\etime{C} = t_3$ and
-$\vtime{A} = \vtime{B} \vtime{C} = 0$, interposed with two
+\end{lemma}
+
+\note{It's possible that these lemmas should be in a different order--
+ we may need \ref{theorem:main} to prove \ref{lemma-rel-etime-vtime+}.}
+
+From these lemmas we can reason about the evaluation time of 
+programs. For example, consider subprograms $A$, $B$, $C$ where 
+$\vtime{A} = \vtime{B} = \vtime{C} = 0$ interposed with two
 sleep statements of duration $s_1$ and $s_2$:
@@
 %%
-Then by the above lemmas, we see that $\etime{eq. \eqref{example:time1}} = 
-s_1 + s_2 + t_3$, iff $t_1 \leq 1$ and $t_2 \leq 2$.
+Then by the above lemmas, we see that if $\etime{A} \leq s_1$ and
+$\etime{B} \leq s_2$ then $\etime{eq. \eqref{example:time1}} = 
+s_1 + s_2 + \etime{C}$.
 
@@
 effectful computation by ignoring the time parameters and 
-providing the usual pure state behaviour: passing the state unchanged.
+providing the usual pure state behaviour of returning the parameter state unchanged
+(named \emph{vT} in this case).
 
@@
   \emph{nowT}, and virtual time \emph{vT}, is the result of evaluating
-  first the left-hand side at time \emph{nowT} and then
-
-right hand side, and then the left hand side.  Note that
-  \ensuremath{\Varid{getCurrentTime}\mathbin{::}\Conid{IO}\;\Conid{Time}} retrieves the time from the operation
-  system.
+  first the left-hand side at time \emph{nowT} and then right-hand side 
+  at the new current time \emph{thenT}.
+
+  The expression \ensuremath{\Varid{getCurrentTime}\mathbin{::}\Conid{IO}\;\Conid{Time}} retrieves the time from
+  the operation system.
 \end{itemize}
 
-The \ensuremath{\Varid{runTime}} operation executes a temporal computation
-inside of the \emph{IO} monad by providing the start time of the
-computation, and with virtual time 0:
+To model the evaluation of a program, the \ensuremath{\Varid{runTime}} operation executes
+a temporal computation inside of the \emph{IO} monad by providing the
+start time of the computation and virtual time 0:
 %%
@@
 ordering and interleaving of calls to the operation system for the
-current time, we consider a program:
+current time, consider the program:
 %%
@@
 \end{hscode}\resethooks
-(where \ensuremath{\Varid{f}\mathrel{=}\Conid{T}\;\Varid{f'},\Varid{g}\mathrel{=}\Conid{T}\;\Varid{g'},\Varid{h}\mathrel{=}\Conid{T}\;\Varid{h'}}) which 
+(where \ensuremath{\Varid{f}\mathrel{=}(\Conid{T}\;\Varid{f'}),\Varid{g}\mathrel{=}(\Conid{T}\;\Varid{g'}),\Varid{h}\mathrel{=}(\Conid{T}\;\Varid{h'})}) which 
 desugars to the following \emph{IO} computation, 
@@
 %
-The above illustrates the repeated calls to \ensuremath{\Varid{getCurrentTime}}, the
+This illustrates the repeated calls to \ensuremath{\Varid{getCurrentTime}}, the
 constant start time parameter, and the threading of virtual time state
@@
 
-Figure~\ref{core-functions} provides a number of effectul operations
-for \emph{Temporal} that access the current time, the start time, get
+Figure~\ref{core-functions} shows a number of effectul operations of 
+ the \emph{Temporal} monad that access the current time, the start time, get
 and set the virtual time, and cause a kernel sleep. These
 are used in the next part of the model.
-
-\begin{figure}[t]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{28}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{time}\mathbin{::}\Conid{Temporal}\;\Conid{Time}{}\<[E]%
-\\
-\>[B]{}\Varid{time}{}\<[7]%
-\>[7]{}\mathrel{=}\Conid{T}\;(\lambda (\anonymous ,\Varid{nowT})\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{nowT},\Varid{vT})){}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{start}\mathbin{::}\Conid{Temporal}\;\Conid{Time}{}\<[E]%
-\\
-\>[B]{}\Varid{start}\mathrel{=}\Conid{T}\;(\lambda (\Varid{startT},\anonymous )\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{startT},\Varid{vT})){}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{getVirtualTime}\mathbin{::}\Conid{Temporal}\;\Conid{VTime}{}\<[E]%
-\\
-\>[B]{}\Varid{getVirtualTime}\mathrel{=}\Conid{T}\;(\lambda (\anonymous ,\anonymous )\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{vT},\Varid{vT})){}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{setVirtualTime}\mathbin{::}\Conid{VTime}\to \Conid{Temporal}\;(){}\<[E]%
-\\
-\>[B]{}\Varid{setVirtualTime}\;\Varid{vT}\mathrel{=}\Conid{T}\;(\lambda \anonymous \to \lambda \anonymous \to \Varid{return}\;((),\Varid{vT})){}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{kernelSleep}\mathbin{::}\Conid{RealFrac}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{Temporal}\;(){}\<[E]%
-\\
-\>[B]{}\Varid{kernelSleep}\;\Varid{t}\mathrel{=}{}\<[18]%
-\>[18]{}\Conid{T}\;(\lambda (\anonymous ,\anonymous )\to \lambda \Varid{vT}\to {}\<[E]%
-\\
-\>[18]{}\hsindent{6}{}\<[24]%
-\>[24]{}\mathbf{do}\;{}\<[28]%
-\>[28]{}\Varid{threadDelay}\;(\Varid{round}\;(\Varid{t}\mathbin{*}\mathrm{1000000})){}\<[E]%
-\\
-\>[28]{}\Varid{return}\;((),\Varid{vT})){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Simple \emph{Temporal} computations, used  by the model}
-\label{core-functions}
-\end{figure}
 
@@
 \column{20}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{30}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{37}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \\
-\>[20]{}\mathbf{if}\;(\Varid{vT'}\mathbin{<}\Varid{diffT})\;\mathbf{then}\;\Varid{return}\;(){}\<[E]%
-\\
-\>[20]{}\hsindent{17}{}\<[37]%
-\>[37]{}\mathbf{else}\;\Varid{kernelSleep}\;(\Varid{vT'}\mathbin{-}\Varid{diffT}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-%
-Thus, \emph{sleep} proceeds by getting the current time, calculating
-the new virtual time \emph{vT'} and updating the virtual time
-state. If the new virtual time is less than the elapsed time
-\emph{diffT} then no actual (kernel) sleeping happens. However, if the
-new virtual time is ahead of the elapsed time, then the process sleeps
-for the difference.
+\>[20]{}\mathbf{if}\;(\Varid{vT'}\mathbin{<}\Varid{diffT}){}\<[E]%
+\\
+\>[20]{}\hsindent{2}{}\<[22]%
+\>[22]{}\mathbf{then}\;\Varid{return}\;(){}\<[E]%
+\\
+\>[20]{}\hsindent{2}{}\<[22]%
+\>[22]{}\mathbf{else}\;\Varid{kernelSleep}\;(\Varid{vT'}\mathbin{-}\Varid{diffT}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%
+where \emph{sleep} proceeds first by getting the current time
+\emph{nowT}, calculating the new virtual time \emph{vT'} and updating
+the virtual time state. If the new virtual time is less than the
+elapsed time \emph{diffT} since the start then no actual (kernel)
+sleeping happens. However, if the new virtual time is ahead of the
+elapsed time, then the process waits for the difference such that the
+elapsed time equals the virtual time.
+
+\begin{figure}[t]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{28}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{time}\mathbin{::}\Conid{Temporal}\;\Conid{Time}{}\<[E]%
+\\
+\>[B]{}\Varid{time}{}\<[7]%
+\>[7]{}\mathrel{=}\Conid{T}\;(\lambda (\anonymous ,\Varid{nowT})\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{nowT},\Varid{vT})){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{start}\mathbin{::}\Conid{Temporal}\;\Conid{Time}{}\<[E]%
+\\
+\>[B]{}\Varid{start}\mathrel{=}\Conid{T}\;(\lambda (\Varid{startT},\anonymous )\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{startT},\Varid{vT})){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{getVirtualTime}\mathbin{::}\Conid{Temporal}\;\Conid{VTime}{}\<[E]%
+\\
+\>[B]{}\Varid{getVirtualTime}\mathrel{=}\Conid{T}\;(\lambda (\anonymous ,\anonymous )\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{vT},\Varid{vT})){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{setVirtualTime}\mathbin{::}\Conid{VTime}\to \Conid{Temporal}\;(){}\<[E]%
+\\
+\>[B]{}\Varid{setVirtualTime}\;\Varid{vT}\mathrel{=}\Conid{T}\;(\lambda \anonymous \to \lambda \anonymous \to \Varid{return}\;((),\Varid{vT})){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{kernelSleep}\mathbin{::}\Conid{RealFrac}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{Temporal}\;(){}\<[E]%
+\\
+\>[B]{}\Varid{kernelSleep}\;\Varid{t}\mathrel{=}{}\<[18]%
+\>[18]{}\Conid{T}\;(\lambda (\anonymous ,\anonymous )\to \lambda \Varid{vT}\to {}\<[E]%
+\\
+\>[18]{}\hsindent{6}{}\<[24]%
+\>[24]{}\mathbf{do}\;{}\<[28]%
+\>[28]{}\Varid{threadDelay}\;(\Varid{round}\;(\Varid{t}\mathbin{*}\mathrm{1000000})){}\<[E]%
+\\
+\>[28]{}\Varid{return}\;((),\Varid{vT})){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\caption{Simple \emph{Temporal} computations, used  by the model}
+\label{core-functions}
+\end{figure}
 
@@
 \column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{16}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \\
-\>[5]{}\mathbf{if}\;(\Varid{vT''}\mathbin{<}\Varid{diffT}){}\<[E]%
-\\
-\>[5]{}\hsindent{2}{}\<[7]%
-\>[7]{}\mathbf{then}\;\Varid{return}\;(){}\<[E]%
-\\
-\>[5]{}\hsindent{2}{}\<[7]%
-\>[7]{}\mathbf{else}\;\Varid{kernelSleep'}\;(\Varid{vT''}\mathbin{-}\Varid{diffT}){}\<[E]%
+\>[5]{}\mathbf{if}\;(\Varid{vT''}\mathbin{<}\Varid{diffT})\;{}\<[24]%
+\>[24]{}\mathbf{then}\;\Varid{return}\;(){}\<[E]%
+\\
+\>[24]{}\mathbf{else}\;\Varid{kernelSleep'}\;(\Varid{vT''}\mathbin{-}\Varid{diffT}){}\<[E]%
 \ColumnHook
@@
 where \ensuremath{\Varid{kernelSleep'}\;\Varid{x}\mathrel{=}\Varid{threadDelay}\;(\Varid{round}\;(\Varid{x}\mathbin{*}\mathrm{1000000}))} is used
-as an alias here (as per the definition of \emph{kernelSleep} in
-Figure~\ref{core-functions}.
+to simplify the code here (as per the definition of \emph{kernelSleep} in
+Figure~\ref{core-functions}).
 
@@
  $\emph{vT''} = \vtime{P} + t$. Therefore, the guard of the 
-\emph{if} expression is $(\vtime{P} + t) < \etime{P}$.
+\ensuremath{\mathbf{if}} expression is $(\vtime{P} + t) < \etime{P}$.
 If the updating of the virtual time state and the computing of 
@@
 
+\note{I suppose this is ok- I'm a bit wary about saying the simplification
+takes less than $\epsilon$. It surely does, but I am only hand waving. 
+We could time $e$ though in the model and show it is less than the schedule
+time. We could go further and time the analogous parts of the SonicPi implementation
+to check that the real $e$ is less than $\epsilon$. This would be good.}
+
 \begin{replemma}{lem:sleep-L}
@@
 
+\note{Rough notes for the proof}
 \begin{proof}
@@
 \end{itemize}
+
+\note{This is actually quite hard. I think I can do it
+by using Lemma 1 and 2 together with some reassociating, but
+haven't time to sort it right now.}
 
@@
 on \emph{Temporal} since all monads are applicative functors and all 
-monads define monoid over $m ()$.
+monads $m$ define a monoid over \ensuremath{\Varid{m}\;()}.
 
@@
 \begin{align*}
-\etime{P; \sleep{} t} = 
+\etime{P; \sleep{} t} \approx 
  \begin{cases}
@@
 
+\note{TODO: report warnings when overrun occurs- should be easy I think
+by adding stuff to sleep.}
+
 \paragraph{Overrun schedule}
+
+\note{TOOD: Again pretty easy, just need some info in the monad
+to say what counts as a bad overrun}
 
@@
 
+\note{TODO}
+
 \paragraph{Acknowledgements}
@@
 \paragraph{Proof} (of Theorem~\ref{theorem:main})
+\note{These are just rough notes}
 Since sequential composition is associative, we can reassociate

### 2014-05-06 paper/paper.tex
s/SonicPi/Sonic Pi and remove all unnecessary whitespace.

@@
 
-\newcommand{\lang}{SonicPi}
+\newcommand{\lang}{Sonic Pi}
 
@@
 
-\note{Introduction to SonicPi}
-
-The underlying programming model of SonicPi provides a way to separate
+\note{Introduction to Sonic Pi}
+
+The underlying programming model of Sonic Pi provides a way to separate
 the ordering of effects from the timing of
@@
 %
-The \lang{} was previously discussed in~\cite{aaron2013sonic}
+\lang{} was previously discussed in~\cite{aaron2013sonic}
 
@@
 We could time $e$ though in the model and show it is less than the schedule
-time. We could go further and time the analogous parts of the SonicPi implementation
+time. We could go further and time the analogous parts of the Sonic Pi implementation
 to check that the real $e$ is less than $\epsilon$. This would be good.}

### 2014-05-06 paper/paper.tex
Add Alan's intro and associated refs

@@
 
+Sonic Pi is a music live coding language that has been designed for
+educational use, as a first programming language~\cite{Aaron2013}. A first
+programming language should be conceptually straightforward and
+syntactically uncluttered. However, it is not straightforward to achieve
+this simplicity in a music live coding language, for reasons largely
+related to the representation of time. Representing musical time in a
+programming language is often problematic, firstly because natural ways
+of describing and organising musical events are multi-threaded (scores,
+chords, resonance, reverb), and secondly because so many standard
+computational formalisms treat execution time as a non-functional
+requirement. Time can be even more problematic in live coding, because
+the creation of the code, as a performance, is interleaved with the
+sound events resulting from its execution. Yet for users of Sonic Pi,
+the creative experience they have, like all experience, arises through
+time - as media theorist Mieke Bal says, "time is where subjectivity is
+produced"~\cite{Bal2002}.
+
+As noted by Julian Rohrhuber~\cite{blackwell_et_al2014}, there have been
+many publications discussing alternative approaches to the
+representation of time in live coding, many having to choose between
+explicit or implicit representation of time, and between description of
+time with reference to internal or external state. These many
+interesting strategies include McLean's formalism of cyclic time in the
+Tidal language~\cite{McLean2013}, and Sorensen's temporal recursion in
+Impromptu/Extempore~\cite{Sorensen2010}. In this paper, we present a formalism that is
+designed to achieve production-quality sound (via the SuperCollider
+synthesis server) while allowing inexperienced programmers in an
+educational setting (typically 11-12 year-old children) to express the
+temporal structure in terms that have an intuitive correspondence to the
+experience and production of musical sounds.
+
 The underlying programming model of Sonic Pi provides a way to separate
@@
 %
-\lang{} was previously discussed in~\cite{aaron2013sonic}
+\lang{} was previously discussed in~\cite{Aaron2013}
 

### 2014-05-07 TemporalExceptions.hs
model with exceptions added

@@ -1,0 +1,116 @@
+{-# LANGUAGE NoMonomorphismRestriction, FlexibleInstances #-}
+
+module TemporalExceptions where
+
+import Data.Time
+import Control.Applicative
+import Control.Concurrent
+import Data.Monoid
+import Debug.Trace
+
+import Temporal (Temporal(..), Time, VTime, diffTime)
+import qualified Temporal as Temporal
+
+data Excp = StrongOverrun VTime | WeakOverrun VTime deriving Show
+
+data TemporalE a = TE (VTime -> Temporal (a, [Excp]))
+
+instance Monad TemporalE where 
+    return a = TE (\_ -> return (a, []))
+    (TE p) >>= q = TE (\eps -> (p eps) >>= (\(a, es) -> do let (TE q') = q a
+                                                           (b, es') <- q' eps
+                                                           return (b, es++es')))
+
+-- Evaluate a temporal computation
+runTime :: VTime -> TemporalE a -> IO (a, [Excp])
+runTime eps (TE c) = do startT <- getCurrentTime
+                        let (T c') = c eps
+                        (y, _) <- c' (startT, startT) 0
+                        return y
+
+
+lift :: Temporal a -> TemporalE a
+lift t = TE (\_ -> fmap (\a -> (a, [])) t)
+
+-- Accessors
+epsilonTime :: TemporalE VTime
+epsilonTime = TE (\eps -> return (eps, []))
+
+time  = lift (Temporal.time)
+start = lift (Temporal.start)
+getVirtualTime = lift (Temporal.getVirtualTime)
+setVirtualTime vT = lift (Temporal.setVirtualTime vT)
+
+-- Lift an IO computation to a temporal computation
+liftIO :: IO a -> TemporalE a
+liftIO k = lift (Temporal.liftIO k)
+
+{-# INLINE kernelSleep #-}
+kernelSleep :: RealFrac a => a -> TemporalE ()
+kernelSleep x = lift (Temporal.kernelSleep x)
+
+
+weakException t = TE (\_ -> return ((), [WeakOverrun t])) >>
+                    (liftIO . putStrLn $ "warning: overran by " ++ (show t))
+strongException t = TE (\_ -> return ((), [StrongOverrun t])) >>
+                    (liftIO . putStrLn $ "WARNING: overran by " ++ (show t))
+
+sleep :: VTime -> TemporalE ()
+sleep delayT = do nowT      <- time
+                  vT        <- getVirtualTime
[…]

### 2014-05-07 paper/paper.tex
small tweaks to early figure

--- 2014-05-01 paper/paper.tex
@@
 
+\CustomVerbatimEnvironment{SVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,xrightmargin=0.5cm,framesep=3mm,commandchars=\\\{\}}
+
 \makeatletter
@@
 \authorinfo{Dominic Orchard}{}{}
-\title{A programming model for temporal coordination (in music)}
+\title{Temporal semantics for a live coding language} 
+% A programming model for temporal coordination (in music)}
 
@@
 \maketitle
+
+\begin{abstract}
+Sonic Pi is a music live coding language that has been designed for
+educational use, as a first programming language. However, it is
+not straightforward to achieve the necessary simplicity of a first language
+in a music live coding setting, for reasons largely related to the
+manipulation of time. The original version of Sonic Pi used a 'sleep'
+function for managing time. However, whilst this approach was conceptually
+simple, it resulted in badly timed music - especially when multiple musical
+threads were executing concurrently. This paper describes an alternative
+programming approach for timing (implemented in Sonic Pi v2.0) which
+maintains syntactic compatibility with version 1, yet provides accurate
+timing via specific interaction between real time and a "virtual time".
+We provide a formal specification of the temporal behaviour of Sonic Pi,
+motivated in relation to other recent approaches to the semantics of time in
+live coding and general computation. We then define a monadic model of the
+Sonic Pi temporal semantics which is sound with respect to this
+specification, using Haskell as a metalanguage. 
+\end{abstract}
 
@@
 
+\begin{SaveVerbatim}{example1}
+play C
+play E
+play G
+sleep 1
+play F
+play A
+play C
+sleep 0.5
+play G
+play B
+play D
+\end{SaveVerbatim}
+
 \begin{figure}[t]
@@
 \begin{minipage}{0.46\linewidth}
-\[
-\hspace{-6em}
-\begin{array}{l}
-\play C \\
-\play E \\ 
-\play G \\
-\sleep 1 \\
-\play F \\
-\play A \\
-\play C \\
-\sleep 0.5 \\
-\play G \\
-\play B \\
-\play D \\
-\end{array}
-\]
+%\[
+%\hspace{-6em}
+%\begin{array}{l}
+%\play C \\
+%\play E \\ 
+%\play G \\
+%\sleep 1 \\
+%\play F \\
+%\play A \\
+%\play C \\
+%\sleep 0.5 \\
+%\play G \\
+%\play B \\
+%\play D \\
+%\end{array}
+%\]
+\BUseVerbatim[fontsize=\footnotesize,baselinestretch=0.97]{example1} \\
 \end{minipage}

### 2014-05-08 paper/paper.tex
some refactoring on exceptions

@@
    \let\\=\@normalcr
-   \(\pboxed}%
+   \let\hspre\(\let\hspost\)%
+   \pboxed}%
   {\endpboxed\)%
@@
 
-\authorinfo{Sam Aaron}{}{}
-\authorinfo{Dominic Orchard}{}{}
+\authorinfo{Sam Aaron}
+           {Computer Laboratory, University of Cambridge, UK}
+           {sam.saaron\ensuremath{\mathord{@}}cl.cam.ac.uk}
+\authorinfo{Dominic Orchard}
+           {Computer Laboratory, University of Cambridge, UK}
+           {dominic.orchard\ensuremath{\mathord{@}}cl.cam.ac.uk}
+\authorinfo{Alan Blackwell}
+           {Computer Laboratory, University of Cambridge, UK}
+           {alan.blackwell\ensuremath{\mathord{@}}cl.cam.ac.uk}
+
 \title{Temporal semantics for a live coding language} 
@@
 \>[3]{}\Varid{return}\;\Varid{a}{}\<[16]%
-\>[16]{}\mathrel{=}\Conid{T}\;(\lambda \anonymous \to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{a},\Varid{vT})){}\<[E]%
+\>[16]{}\mathrel{=}\Conid{T}\;(\mathbin{\char92 \char95 }\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{a},\Varid{vT})){}\<[E]%
 \\[\blanklineskip]%
@@
 \interp{x = P; Q} & = \interp{P} \ensuremath{\bind (\lambda \Varid{x}\to } \interp{Q}) \\
-\interp{P; Q} & = \interp{P} \ensuremath{\bind (\lambda \anonymous \to } \interp{Q}) \\
+\interp{P; Q} & = \interp{P} \ensuremath{\bind (\mathbin{\char92 \char95 }\to } \interp{Q}) \\
 \interp{\sleep e} & = \emph{sleep} \, \interp{e}
@@
 \>[7]{}\mathrel{=}\Conid{T}\;(\lambda (\anonymous ,\Varid{nowT})\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{nowT},\Varid{vT})){}\<[E]%
-\\[\blanklineskip]%
+\\
+\>[B]{}\\[-1.5em]{}\<[E]%
+\\
 \>[B]{}\Varid{start}\mathbin{::}\Conid{Temporal}\;\Conid{Time}{}\<[E]%
@@
 \>[B]{}\Varid{start}\mathrel{=}\Conid{T}\;(\lambda (\Varid{startT},\anonymous )\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{startT},\Varid{vT})){}\<[E]%
-\\[\blanklineskip]%
+\\
+\>[B]{}\\[-1.5em]{}\<[E]%
+\\
 \>[B]{}\Varid{getVirtualTime}\mathbin{::}\Conid{Temporal}\;\Conid{VTime}{}\<[E]%
@@
 \>[B]{}\Varid{getVirtualTime}\mathrel{=}\Conid{T}\;(\lambda (\anonymous ,\anonymous )\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{vT},\Varid{vT})){}\<[E]%
-\\[\blanklineskip]%
+\\
+\>[B]{}\\[-1.5em]{}\<[E]%
+\\
 \>[B]{}\Varid{setVirtualTime}\mathbin{::}\Conid{VTime}\to \Conid{Temporal}\;(){}\<[E]%
 \\
-\>[B]{}\Varid{setVirtualTime}\;\Varid{vT}\mathrel{=}\Conid{T}\;(\lambda \anonymous \to \lambda \anonymous \to \Varid{return}\;((),\Varid{vT})){}\<[E]%
-\\[\blanklineskip]%
+\>[B]{}\Varid{setVirtualTime}\;\Varid{vT}\mathrel{=}\Conid{T}\;(\mathbin{\char92 \char95 }\to \mathbin{\char92 \char95 }\to \Varid{return}\;((),\Varid{vT})){}\<[E]%
+\\
+\>[B]{}\\[-1.5em]{}\<[E]%
+\\
 \>[B]{}\Varid{kernelSleep}\mathbin{::}\Conid{RealFrac}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{Temporal}\;(){}\<[E]%
@@
 
-\subsection{Emitting overrun exceptions}
+\subsection{Emitting overrun warnings}
+
+We extend the \ensuremath{\Conid{Temporal}} monad with an additional parameter for the
+$\epsilon$ time (maximum allowable overrun) and an output stream for
+sending ``warnings'' when overruns occur.
+
+Overrun warnings are either \emph{strong}, when the real time 
+is more than $\epsilon$ ahead of virtual time, or \emph{weak} when the real
+time is less than $\epsilon$ ahead of virtual time. That is:
+%%
+\begin{itemize}
+\item{$\etime{P} > (\vtime{P} + \epsilon) \Rightarrow \interp{P} \leadsto$ \emph{strong warning} }
+\item{$(\vtime{P} + \epsilon) \leq \etime{P} > \vtime{P} \Rightarrow \interp{P} \leadsto$ \emph{weak warning}}
+\end{itemize}
+%%
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{data}\;\Conid{Warning}\mathrel{=}\Conid{StrongOverrun}\;\Conid{VTime}\mid \Conid{WeakOverrun}\;\Conid{VTime}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\mathbf{data}\;\Conid{TemporalE}\;\Varid{a}\mathrel{=}\Conid{TE}\;(\Conid{VTime}\to \Conid{Temporal}\;(\Varid{a},[\mskip1.5mu \Conid{Warning}\mskip1.5mu])){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\mathbf{instance}\;\Conid{Monad}\;\Conid{TemporalE}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{return}\;\Varid{a}\mathrel{=}\Conid{TE}\;(\mathbin{\char92 \char95 }\to \Varid{return}\;(\Varid{a},[\mskip1.5mu \mskip1.5mu])){}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}(\Conid{TE}\;\Varid{p})\bind \Varid{q}\mathrel{=}\Conid{TE}\;(\lambda \Varid{eps}\to \mathbf{do}\;(\Varid{a},\Varid{es})\leftarrow \Varid{p}\;\Varid{eps}{}\<[E]%
+\\
+\>[5]{}\hsindent{30}{}\<[35]%
+\>[35]{}\mathbf{let}\;(\Conid{TE}\;\Varid{q'})\mathrel{=}\Varid{q}\;\Varid{a}{}\<[E]%
+\\
+\>[5]{}\hsindent{30}{}\<[35]%
+\>[35]{}(\Varid{b},\Varid{es'})\leftarrow \Varid{q'}\;\Varid{eps}{}\<[E]%
+\\
+\>[5]{}\hsindent{30}{}\<[35]%
+\>[35]{}\Varid{return}\;(\Varid{b},\Varid{es}\plus \Varid{es'})){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 

### 2014-05-08 TemporalExceptions.hs

@@ -11,18 +11,19 @@ import Debug.Trace
 import Temporal (Temporal(..), Time, VTime, diffTime)
 import qualified Temporal as Temporal
 
-data Excp = StrongOverrun VTime | WeakOverrun VTime deriving Show
+data Warning = StrongOverrun VTime | WeakOverrun VTime deriving Show
 
-data TemporalE a = TE (VTime -> Temporal (a, [Excp]))
+data TemporalE a = TE (VTime -> Temporal (a, [Warning]))
 
 instance Monad TemporalE where 
     return a = TE (\_ -> return (a, []))
-    (TE p) >>= q = TE (\eps -> (p eps) >>= (\(a, es) -> do let (TE q') = q a
-                                                           (b, es') <- q' eps
-                                                           return (b, es++es')))
+    (TE p) >>= q = TE (\eps -> do (a, es) <- p eps
+                                  let (TE q') = q a
+                                  (b, es') <- q' eps
+                                  return (b, es++es'))
 
 -- Evaluate a temporal computation
-runTime :: VTime -> TemporalE a -> IO (a, [Excp])
+runTime :: VTime -> TemporalE a -> IO (a, [Warning])
 runTime eps (TE c) = do startT <- getCurrentTime
                         let (T c') = c eps
                         (y, _) <- c' (startT, startT) 0
@@ -64,8 +65,8 @@ sleep delayT = do nowT      <- time
                   eps       <- epsilonTime
                   let diffT = diffTime nowT startT
                   if (vT' < diffT) then if ((vT' + eps) < diffT) 
-                                        then weakException (diffT - vT')
-                                        else strongException (diffT - vT')
+                                        then strongException (diffT - vT')
+                                        else weakException (diffT - vT')
                                    else kernelSleep (vT' - diffT)
 
 duration :: TemporalE b -> TemporalE (b, VTime)

### 2014-05-08 paper/paper.tex
started on overrun text

@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{data}\;\Conid{Warning}\mathrel{=}\Conid{StrongOverrun}\;\Conid{VTime}\mid \Conid{WeakOverrun}\;\Conid{VTime}{}\<[E]%
+\>[B]{}\mathbf{data}\;\Conid{Warning}\mathrel{=}\Conid{Strong}\;\Conid{VTime}\mid \Conid{Weak}\;\Conid{VTime}{}\<[E]%
 \\[\blanklineskip]%
-\>[B]{}\mathbf{data}\;\Conid{TemporalE}\;\Varid{a}\mathrel{=}\Conid{TE}\;(\Conid{VTime}\to \Conid{Temporal}\;(\Varid{a},[\mskip1.5mu \Conid{Warning}\mskip1.5mu])){}\<[E]%
+\>[B]{}\mathbf{data}\;\Conid{TemporalE}\;\Varid{a}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{TE}\;(\Conid{VTime}\to \Conid{Temporal}\;(\Varid{a},[\mskip1.5mu \Conid{Warning}\mskip1.5mu])){}\<[E]%
 \\[\blanklineskip]%

### 2014-05-08 paper/paper.tex
started discussion on laws

@@
    \let\\=\@normalcr
-   \let\hspre\(\let\hspost\)%
-   \pboxed}%
+   \(\pboxed}%
   {\endpboxed\)%
@@
 \>[3]{}\Varid{return}\;\Varid{a}{}\<[16]%
-\>[16]{}\mathrel{=}\Conid{T}\;(\mathbin{\char92 \char95 }\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{a},\Varid{vT})){}\<[E]%
+\>[16]{}\mathrel{=}\Conid{T}\;(\lambda \anonymous \to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{a},\Varid{vT})){}\<[E]%
 \\[\blanklineskip]%
@@
 \interp{x = P; Q} & = \interp{P} \ensuremath{\bind (\lambda \Varid{x}\to } \interp{Q}) \\
-\interp{P; Q} & = \interp{P} \ensuremath{\bind (\mathbin{\char92 \char95 }\to } \interp{Q}) \\
+\interp{P; Q} & = \interp{P} \ensuremath{\bind (\lambda \anonymous \to } \interp{Q}) \\
 \interp{\sleep e} & = \emph{sleep} \, \interp{e}
@@
 \\
-\>[B]{}\Varid{setVirtualTime}\;\Varid{vT}\mathrel{=}\Conid{T}\;(\mathbin{\char92 \char95 }\to \mathbin{\char92 \char95 }\to \Varid{return}\;((),\Varid{vT})){}\<[E]%
+\>[B]{}\Varid{setVirtualTime}\;\Varid{vT}\mathrel{=}\Conid{T}\;(\lambda \anonymous \to \lambda \anonymous \to \Varid{return}\;((),\Varid{vT})){}\<[E]%
 \\
@@
 
+The \ensuremath{\Conid{Temporal}} moand is ``weak'', in the sense that the standard monad
+laws do not hold in general.  For example, one of the unit laws is that:
+%%
+\begin{equation}
+%|(return x) >>= (\y -> f y)| \equiv |f x|
+m >>= return \equiv m
+\label{law-example}
+\end{equation}
+%%
+where \ensuremath{\Varid{m}\mathbin{::}\Conid{Temporal}\;\Varid{a}}.  However, for \ensuremath{\Conid{Temporal}} this law is
+violated in cases where \ensuremath{\Varid{m}} depends on the current time. For example, let
+\ensuremath{\Varid{m}} be defined:
+%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{m}\mathrel{=}\mathbf{do}\;{}\<[10]%
+\>[10]{}\Varid{kernelSleep}\;\mathrm{1}{}\<[E]%
+\\
+\>[10]{}\Varid{start}\leftarrow \Varid{start}{}\<[E]%
+\\
+\>[10]{}\Varid{end}\leftarrow \Varid{time}{}\<[E]%
+\\
+\>[10]{}\Varid{return}\;(\Varid{diffTime}\;\Varid{end}\;\Varid{start})\mbox{\onelinecomment  duration of computation}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%%
+Then we can run an experiment in GHCi to see that different results are possible:
+%%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbin{*}\Conid{Main}\mathbin{>}\Varid{runTime}\mathbin{\$}\Varid{m}\bind \Varid{return}{}\<[E]%
+\\
+\>[B]{}\mathrm{1.001113}\;\Varid{s}{}\<[E]%
+\\
+\>[B]{}\mathbin{*}\Conid{Main}\mathbin{>}\Varid{runTime}\mathbin{\$}\Varid{m}{}\<[E]%
+\\
+\>[B]{}\mathrm{1.00114}\;\Varid{s}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+% The difference in results follows from the additional reduction
+required on \ensuremath{(\bind )} in the first case (left-hand side of
+\eqref{law-example}).
+
 \paragraph{Quotienting by non-time dependent functions}
@@
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{return}\;\Varid{a}\mathrel{=}\Conid{TE}\;(\mathbin{\char92 \char95 }\to \Varid{return}\;(\Varid{a},[\mskip1.5mu \mskip1.5mu])){}\<[E]%
+\>[5]{}\Varid{return}\;\Varid{a}\mathrel{=}\Conid{TE}\;(\lambda \anonymous \to \Varid{return}\;(\Varid{a},[\mskip1.5mu \mskip1.5mu])){}\<[E]%
 \\

### 2014-05-08 Tests.hs

@@ -1,4 +1,6 @@
 import Temporal
+import Control.Applicative 
+import Data.Time
 
 tests slp = do putStrLn "test 1, expected = 2s, (2s, 1s, 3s)" 
                r1 <- getMean 5 (test1 slp)
@@ -40,9 +42,9 @@ test1 slp =
         b <- time
         slp 1
         c <- time
-        return $ (diffUTCTime b a, -- 2s
-                  diffUTCTime c b, -- 1s
-                  diffUTCTime c a) -- 3s
+        return $ (diffTime b a, -- 2s
+                  diffTime c b, -- 1s
+                  diffTime c a) -- 3s
 
 -- [test2]_t = 3
 test2 slp =
@@ -51,9 +53,9 @@ test2 slp =
          b <- time
          slp 2
          c <- time
-         return $ (diffUTCTime b a, -- 1s
-                   diffUTCTime c b, -- 2s
-                   diffUTCTime c a) -- 3s
+         return $ (diffTime b a, -- 1s
+                   diffTime c b, -- 2s
+                   diffTime c a) -- 3s
 
 -- [test3A]_t = 6
 test3A slp = (kernelSleep 2) *> (slp 1) *> (slp 2) *> (kernelSleep 3)
@@ -80,7 +82,7 @@ test4A slp = do slp 2
                 s <- start
                 e <- time
                 r <- getVirtualTime
-                return $ (diffUTCTime e s, r)
+                return $ (diffTime e s, r)
 
 -- [test4B]_t = 5
 test4B slp = do kernelSleep 5
@@ -88,7 +90,7 @@ test4B slp = do kernelSleep 5
                 s <- start
                 e <- time
                 r <- getVirtualTime
-                return $ (diffUTCTime e s, r)
+                return $ (diffTime e s, r)
 
 getMean :: (Fractional b) => Integer -> Temporal b -> IO (b, NominalDiffTime)
 getMean n k = do (r, d) <- getMean' n
@@ -135,9 +137,11 @@ assocLaw f g m = do left  <- runTime $ do { y <- do { x <- m; f x; }; g y }
                     putStrLn $ show right
 
 
-law1Foo = unitLaw (\t -> do { sleep t; time; }) 1.0
-law2Foo = unitLaw2 (T (\(t, t') _ -> return $ (diffUTCTime t' t, 0)))
[…]

### 2014-05-08 paper/paper.tex
some work on monad laws and warnings

@@
 \end{hscode}\resethooks
-% The difference in results follows from the additional reduction
-required on \ensuremath{(\bind )} in the first case (left-hand side of
-\eqref{law-example}).
+(note, these results are also non-deterministic).  The difference in
+results follows from the additional reduction required on \ensuremath{(\bind )} in
+the first case (left-hand side of \eqref{law-example}). Note that we
+calculate a duration here since using the absolute time produced by
+\ensuremath{\Varid{time}} would be disingenuous, since we are evaluating \ensuremath{\Varid{m}\bind \Varid{return}}
+and \ensuremath{\Varid{m}} at different start times.
+
+\paragraph{Laws with respect to $\epsilon$}
 
